---
title: "Dispersal constraints shape the sign and magnitude of biodiversity responses to forest loss and fragmentation"
author: ""
output:
  bookdown::word_document2:
    reference_docx: resources/refword.docx   # relativo ao Rmd, como o .bib
    number_sections: false
    fig_caption: true
  bookdown::pdf_document2:
    latex_engine: lualatex
    pandoc_args: [
      "--number-lines",
      "--variable", "geometry=a4paper",
      "--variable", "geometry=twoside",
      "--variable", "geometry=top=3cm",
      "--variable", "geometry=bottom=2cm",
      "--variable", "geometry=inner=3cm",
      "--variable", "geometry=outer=2cm"
      ]
  pdf_document: default
  html_document:
    df_print: paged
bibliography: resources/citations2.bib
csl: "resources/ecography.csl"
editor_options:
  chunk_output_type: console
header-includes:
  - \usepackage{polyglossia}
  - \setmainlanguage{english}
  - \renewcommand{\contentsname}{Table of Contents}
  - \usepackage{caption}
  - \captionsetup[figure]{name=Figure, font=it}
  - \captionsetup[table]{name=Table, font=it}
  - \usepackage{tcolorbox}
  - \usepackage{hyperref}
  - \hypersetup{colorlinks=true, linkcolor=blue, citecolor=red, urlcolor=blue}
  - \usepackage[brazilian,hyperpageref]{backref}
  - \tcbuselibrary{breakable}
  - \tcbuselibrary{skins}   
  - \usepackage{lmodern}
  - \usepackage{fontspec}
  - \usepackage{threeparttable} 
  - \usepackage{caption}
  - \usepackage{setspace}
  - \usepackage{etoolbox}
  - \setmainfont{Times New Roman}
  - \fontsize{12}{22}
  - \setlength{\parindent}{1.5cm}
  - \onehalfspacing
  - \makeatletter
  - \patchcmd{\@startsection}{\@afterindentfalse}{\@afterindenttrue}{}{}
  - \makeatother
  - \captionsetup[figure]{font=it}
  - \captionsetup[table]{font=it}
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup resultados,include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = FALSE,eval = TRUE,message = FALSE,warning = FALSE,cache=FALSE,
                      fig.pos = "H", fig.show = "hold",
                      dpi = 120)
library(gridExtra)
library(ggplot2)
library(readr)
library(dplyr)
library(plyr)
library(flextable)
library(officer)
library(magick)
library(glue)
```

```{r trackdown compilacao integral, eval=FALSE, include=FALSE}
library(trackdown)
# both overwrite: 
# overwrite the current google drive file with the current Rmd:
upload_file(file="./1_to_compile_dissertacao_EM_USO/compilacao_integral.Rmd",
            gpath="mestrado/artigo_principal/secoes_texto/",
            hide_code=TRUE)
update_file(file="./1_to_compile_dissertacao_EM_USO/compilacao_integral.Rmd",
            gpath = "mestrado/artigo_principal/secoes_texto/",
            hide_code = FALSE)
# overwrite the current Rmd with the current google drive 
download_file(file="compilacao_integral.Rmd",
              gpath = "mestrado/artigo_principal/secoes_texto/")
```

 
# Abstract {-}

Habitat loss and fragmentation are major drivers of biodiversity change in tropical forests, yet their independent effects remain inconsistent across systems. This inconsistency may arise because landscape effects are typically inferred from configuration alone, leaving functional connectivity, and the dispersal processes that generate it, implicit. Here we ask how dispersal limitation modulates the effects of forest loss and fragmentation per se on local tree diversity. We used a spatially explicit neutral model (SENM) calibrated against 105 Atlantic Forest tree inventories, and constructed three experimental landscapes per plot: contemporary fragmented, pristine continuous, and clumped (remaining forest aggregated into a single block). We define vulnerability operationally as the dependence of local richness on external species input (U) required, under SENM assumptions, to sustain observed richness; log(Ui/Uj) measures how this dependence changes between landscapes. These contrasts isolated the total effect of forest loss and fragmentation, fragmentation per se, and forest cover loss, while making connectivity an emergent property of dispersal across landscape configuration. Model congruence was high under moderate-to-strong dispersal limitation but declined under weak limitation, restricting inference to 66 plots. Forest cover loss dominated vulnerability in low-cover landscapes, with the strongest effects at both extremes of the dispersal range. Fragmentation per se was weaker and more variable, depending on how configuration interacted with dispersal limitation and local community properties rather than on configuration alone. By making the counterfactual contrasts and their underlying demographic mechanisms explicit, the framework treats variation among plots as part of the landscape effect to be explained rather than averaged out. This provides a way to reconcile general landscape heuristics with context-specific mechanistic explanations.

# Key-words{-}

# Introduction (371 palavras)

Habitat loss and fragmentation are major drivers of biodiversity change in tropical forests (@diaz_pervasive_2019). Yet empirical estimates of the independent effect of fragmentation remain highly inconsistent, ranging from negative to neutral or even positive across systems (@fletcher2018habitat, @fahrig2019habitatA, @miller2019does, @valente2023toward). Part of this uncertainty may arise because landscape effects are often inferred from landscape configuration alone, despite connectivity being widely recognized as a fundamental process underlying biodiversity responses (@villard_beyond_2014; @hanski_2015; @rivaEbanskleite2024). Resolving this gap is not only a theoretical challenge but also a practical necessity for conservation planning, in which different traditions (from generalizable heuristics to context-specific mechanistic models) offer contrasting yet potentially complementary paths forward (@arroyo-rodriguez_designing_2020; @banks-leite_suggestion_2021; @arroyo-rodriguez_preserving_2021). 

Functional connectivity is the movement of individuals through remaining habitat and emerges from the interaction between landscape structure and species dispersal ability (@auffret2017plant; @hatfield2018using). By regulating the exchange of individuals among local communities, connectivity modifies the demographic balance between local extinction and recolonization (@kindlmann2008connectivity; @pardini_beyond_2010; @may_dispersal_2012 @baguette2013individual; @campos2013; @chase_biodiversity_2020). Thus, for a given landscape configuration, different levels of dispersal limitation produce different connectivity regimes and thereby alter colonization, local replacement, and extinction dynamics. Understanding landscape effects therefore requires evaluating how variation in dispersal modifies the demographic consequences of landscape structure.

However, empirical studies typically estimate landscape effects by correlating structural landscape metrics with local biodiversity patterns (@frazier2017). This leaves the role of connectivity implicit, primarily because species dispersal is rarely known in real landscapes (@baguette2013individual; @bullock_synthesis_2017). As a result, there is little consensus on which configuration metrics best represent connectivity or which spatial extents are most relevant for estimating biodiversity responses to landscape change (@fletcher2018habitat; @fahrig2019habitatA; @miller2019does; @valente2023toward; @fletcher_jr_addressing_2023). Consequently, current approaches often infer ecological responses from landscape structure without explicitly representing the demographic processes through which functional connectivity regulates the maintenance of local biodiversity. 

Here we propose a framework to infer plot‑specific landscape effects on local tree diversity using a spatially explicit neutral model calibrated against empirical tree communities. By constructing counterfactual landscapes that experimentally decouple habitat loss from spatial configuration, we estimate the independent and combined effects of forest loss and fragmentation while explicitly accounting for dispersal limitation. Specifically, we ask how variation in dispersal limitation affects the magnitude and direction of these landscape effects along an empirical forest‑cover gradient.
	 	
# Materials and Methods (971 palavras)



## Study system and empirical data



We used the Atlantic Forest as a model system for investigating fragmentation effects in tropical forests. This biome retains less than 30% of its original forest cover, distributed as a mosaic that provides a broad empirical gradient of forest cover and spatial configuration (@vancine2024atlantic). Tree community data were obtained from the TreeCo database (@de_lima_how_2015; @de2020erosion), selecting inventories that: (i) sampled trees with DBH ≥ 4.8 cm; (ii) used single plots ≥ 1 ha; and (iii) had high taxonomic reliability and precise coordinates. Landscape structure was characterized using MapBiomas Collection 6 land cover maps (@mapbiomas), with forest cover defined as native forest formations including secondary stages (Supplementary material Appendix 1). For each plot, we extracted a 16 × 16 km landscape window centered on the plot location, chosen to exceed typical buffers and ensure stabilization of model dynamics (Supplementary material Appendix 2).

## Experimental landscape design
To disentangle habitat loss from fragmentation per se, we constructed three experimental landscapes for each plot (Fig. SI1): (i) the **contemporary fragmented landscape** (observed cover and configuration); (ii) a **pristine continuous-forest landscape** (no habitat loss); and (iii) a **clumped landscape** in which all remaining forest was spatially aggregated into a single block centered on the plot, preserving total habitat amount while minimizing fragmentation. Comparing outcomes across these landscapes isolated three effects: FLF, the total effect of forest loss and fragmentation (pristine vs. fragmented); FPS, fragmentation per se (clumped vs. fragmented); and LFC, forest cover loss (pristine vs. clumped). This design resolves a key inferential problem: because habitat removal inevitably alters configuration, complete orthogonality between amount and fragmentation is only achievable through experimental manipulation (@didham_rethinking_2012; @fahrig_resolving_2022).

## Spatially explicit neutral model
We used a Spatially Explicit Neutral Model (SENM; @rosindell2008) as a mechanistic framework linking landscape structure to local diversity. The SENM assumes ecological equivalence among individuals, with community dynamics emerging from stochastic birth, death, dispersal, and external species input. Two parameters are central in our design: **U**, the probability that a lineage originates from a newly introduced species (external input), and **k**, which quantifies dispersal limitation as the expected proportion of propagules retained near parental individuals. Under this formulation, connectivity emerges from the interaction between demographic dynamics in space and landscape configuration, making the SENM particularly suited for investigating how dispersal limitation modulates biodiversity responses to landscape change (@campos2012; @azaele2016; @thompsonPycoalescence2020).

## Calibration, validation, and vulnerability metric
We calibrated the SENM independently for each plot, dispersal scenario and experimental landscape, adjusting U so that simulated richness matched observed richness (semi-analytical method; @rosindell2008). Because dispersal is unknown for the communities studied, we explored 20 dispersal scenarios (k = 0.99 to 0.05), where k is the proportion of propagules landing within the distance to the nearest neighbour. For each scenario, the spatial extent was defined in pristine landscapes, where U is affected only by dispersal and extent, as the smallest extent accumulating 75% of the total reduction in U across extents (0.5–16 km); we assumed this extent to be sufficient for the other experimental landscapes (Supplementary material Appendix 2). Calibration, extent selection and validation against observed SADs covered the full range of scenarios, which allowed us to characterise where the model is adequate. Landscape effects, however, were estimated and interpreted only for 0.50 ≤ k ≤ 0.99, scenarios in which at least half of the propagules remain within the immediate neighbourhood of the parent tree [dispersal distance of X–Y m at the observed density], consistent with empirical dispersal kernels of trees (@bullock_synthesis_2017; @clark1999seed; Appendix 1 and 2).
Before interpreting effects, we validated the SENM by testing whether simulated communities reproduced empirical species abundance distributions (SADs) using Kolmogorov-Smirnov tests with bootstrap resampling (@dowd_twosamples_2023). We retained only simulation batches in which ≥75% of replicate SADs were congruent across all three landscapes, and restricted interpretation to dispersal regimes with consistently high mean congruence (Supplementary material Appendix 3).
Within the SENM, U measures the dependence of local diversity on external sources required to maintain richness under drift (@rosindell2008, @condit2012). Differences in U between landscapes therefore provide a mechanistic measure of vulnerability. For each plot and regime, we calculated log-ratios of U:
- **FLF** (total effect) = log(U_pristine / U_fragmented)
- **FPS** (fragmentation per se) = log(U_clumped / U_fragmented)
- **LFC** (forest cover loss) = log(U_pristine / U_clumped)
These contrasts decompose the total effect additively: FLF = FPS + LFC.

## Statistical analyses
We analysed variation in SENM congruence (the number of congruent SADs among 100 replicates) with binomial hierarchical generalised additive models (HGAMs; @Pedersen2019). In this framework (@wood2017), s() denotes a smooth, nonlinear function of one variable, estimated with a penalised spline; te() denotes a tensor-product smooth of two variables, which allows the shape of the response to one variable to change with the other (an interaction), in contrast to additive smooths; and plot-level random smooths (written s(k)|Plot in tables) allow the response to k to differ among plots. The most complex congruence model included landscape-specific smooths of k, a plot-level random smooth of k per landscape, and a bivariate smooth of plot coordinates; simpler models removed these terms progressively.


For each landscape effect (FLF, FPS, LFC; log-ratios of U), we fitted Gaussian HGAMs as a function of k and remaining forest cover (%FC). Candidate models comprised (i) plot-level intercepts only, (ii) a smooth of k, (iii) additive smooths of k and %FC, and (iv) a tensor-product smooth of %FC and k, each combined with either a plot-level intercept or a plot-level random smooth of k. Models were fitted by restricted maximum likelihood (REML) with the R package mgcv (@wood2017) and compared using AICc, ΔAICc and Akaike weights; we also report deviance explained and effective degrees of freedom (edf) as measures of fit and complexity. For the congruence models, we also tested residual spatial autocorrelation with Moran's I on plot-mean residuals. Model formulas and diagnostics are provided in the Supporting Information 3.



# Results (249+299+433=981)

<p> <!--Resultados auxiliares: exposição e citação do SI -->
A total of 107 forest inventories from the TreeCo database met the initial selection criteria (Table SI1). The final dataset comprised 105 plots after excluding two sites where the fragmented landscape configuration prevented reconstruction of the sampled plot as a square (Fig. SI1, Tab. SI1). The dataset spans a broad geographic range across the Atlantic Forest and encompasses substantial variation in abundance (median = 1,528 individuals, IQR = 608, range = [649 ; 12,105], species richness (median = 107 species, IQR = 66, range = [26 ; 458] and forest cover at the landscape level (median = 60.94%, IQR = 63.63%, range = [3.33% ; 100%]. Landscape maps for all study sites are shown in Figures SI3.3.1.1-SI3.3.1.2.
</p>

<p> <!--Calibração e sensibilidade: escala suficiente -->
Across dispersal scenarios and plots, U estimates from pristine landscape simulations at the largest spatial scale varied substantially (Figs. SI2.2, SI2.3, SI2.8). Applying a 75% threshold of the total reduction in U yielded more consistent estimates of the sufficient spatial scale (Fig. SI2.7). Under moderate dispersal limitation (k ≈ 0.70), uncertainty increased because differences among spatial scales became smaller than the variation among simulation replicates for a given combination of k and plot (Figs. SI2.4:SI2.6, SI2.9). Linear interpolation within the 75% threshold indicated a sufficient spatial scale of 2 km for scenarios with k ≥ 0.35. For k < 0.30, the sufficient spatial scale increased to 4 km.
</p>


## Dispersal limitation constrains the empirical validity of a spatially explicit neutral model {Congruencia}

<p> <!--Descrição do padrão empírico chave para o critério post-hoc -->
Mean congruence between simulated and observed SADs exceeded 75% across most dispersal scenarios, but declined under very weak dispersal limitation, particularly in clumped and pristine landscapes (Fig. 1). In 103 plots (98.1%), there was at least one parameter setting (dispersal and landscape) in which the 75% congruence criterion was met. 

Congruence varied substantially among plots, with non-linear responses to the interaction between dispersal limitation and landscape configuration. Accordingly, the best-supported model included a plot-level interaction between dispersal and landscape configuration, together with a spatial effect accounting for proximity among plots (Appendix S3).
</p>


<p> <!--a importância da interação entre k e landscape-->
Although the full model received the strongest support, a simpler model excluding spatial coordinates explained the same proportion of deviance (94%), showed no residual spatial autocorrelation (Fig. SI3.1), and had moderately lower support (ΔAICc = 7.16; Table SI3.2). In contrast, models that omitted the plot-level interaction exhibited significant positive spatial autocorrelation and a substantial drop in explained deviance (64% and 59%; Table SI3.2). These results indicate that the plot-level interaction captured the dominant congruence patterns, with spatial proximity providing only a refinement.
</p>

<p> <!--Critério post-hoc-->
Congruence declined as dispersal limitation weakened (Fig. 1), indicating that SENM better reproduces observed SADs under moderate-to-strong dispersal limitation. Subsequent analyses were restricted to scenarios with k ≥ 0.50, where mean congruence consistently exceeded 75% across all experimental landscapes. Applying this criterion together with the predefined requirement of at least 75% congruence across all three landscapes yielded 66 selected plots (62.8%) for the vulnerability analyses. Selected and excluded plots exhibited similar environmental characteristics and showed no evidence of spatial segregation (Appendix SI3).
</p>


## Forest cover and dispersal jointly shape vulnerability to forest loss and fragmentation

<p>
Differences in U between landscapes reflect variation in the external replenishment required to sustain observed diversity patterns. Accordingly, log(Ui/Uj) quantifies landscape‑induced vulnerability across three experimental contrasts: total effect of forest loss and fragmentation (FLF; pristine vs. fragmented), fragmentation per se (FPS; clumped vs. fragmented), and forest cover loss (LFC; pristine vs. clumped). In fig. 2, plots with greater than or equal 99% forest cover were excluded from the graphical display; only the amplitude (maximum = 0.075; minimum = -0.074) and the central 90% interval (95th quantile = 0.04; 5th quantile = -0.038) of log Ui/Uj are indicated on the y-axis. This variability reflects the stochasticity inherent to the spatially explicit neutral model simulation.
</p>

<p>
Model selection indicated that the interaction between remaining forest cover (%FC) and dispersal limitation (k) improved model fit for FLF and LFC, but not for FPS (Tab. 1). For FPS, the best-supported model included only additive effects of %FC and k. Notably, across all three landscape effects, accounting for plot-specific dispersal responses was essential to achieve high explanatory power (>75% deviance explained); removing this component reduced deviance explained by 25 to 40 percentage points (Tab. 1).
</p>

<p>
The LFC effect became increasingly negative in plots with reduced %FC and was negligible in plots with high %FC (Fig. 2). Across dispersal regimes, LFC exhibited a non-linear response to dispersal limitation (k), with minimum vulnerability at intermediate values (k ≈ 0.70) and stronger effects under both strong and weak dispersal limitations. This pattern indicates that habitat loss exerts its strongest influence when dispersal limitation is either strong or weak, whereas intermediate dispersal partially buffers vulnerability. 
</p>

<p>
The FPS effect was generally weak, with most values clustering around the null reference in landscapes with forest cover above 60%, and becoming weakly negative at low forest cover (<30%) (Fig. 2). Only a few plots showed plot-specific patterns, with some displaying pronounced negative effects under moderate-to-mild dispersal limitation and others showing weak positive effects under severe limitation. Although FPS occasionally increased in plots with reduced %FC, the most pronounced effects occurred in a small number of plots with intermediate %FC. Thus, fragmentation per se acts primarily as a context-dependent modifier of vulnerability rather than a dominant driver.
</p>


<p>
The FLF effect became more negative as %FC decreased, although a few plots displayed patterns resembling FPS at intermediate-to-high %FC (Fig. 2). Overall, FLF exhibited greater variability, integrating features of both LFC and FPS. 
</p>


<p>
A complementary analysis showed that differences in magnitude between LFC and FPS were generally small and often fell within the null reference range, although LFC tended to dominate in plots with reduced %FC (Fig. SI4.1). Across plots, LFC and FPS frequently differed in sign (41%-65%). The proportion of cases in which LFC and FPS differed in sign increased with %FC. When both effects were negative, the total FLF was correspondingly more negative. In contrast, when the effects differed in sign, positive FPS partially counterbalanced negative LFC. 
</p>

<p>
These contrasting signs were most evident in low-cover landscapes, where LFC dominated, and in a small number of intermediate-cover plots, where FPS effects prevailed (Fig. 2; Fig. S4.1). Together, these patterns indicate that the combined effect of forest loss and fragmentation is predominantly negative in low-cover landscapes, while in higher-cover landscapes it may reflect plot-specific effects of spatial configuration.
<p>

# Discussion (2373)

<p>
Functional connectivity is widely recognized as the mechanism linking landscape structure to biodiversity responses, yet its demographic basis is rarely represented explicitly (@swift2010critical; @villard_beyond_2014; @rivaEbanskleite2024). In the SENM, connectivity emerges from individual dispersal across a spatially explicit landscape (@rosindell2008; @campos2012, @campos2013; @thompsonPycoalescence2020). Model congruence was high across a broad range of dispersal regimes and experimental landscapes, although it varied substantially among plots. For plots showing consistently high congruence under moderate-to-severe dispersal limitation, landscape effects could be estimated by contrasting experimental landscapes at the plot level. These analyses revealed that the signal and magnitude of forest loss and fragmentation depend on the landscape-dispersal interplay, a finding that helps explain under which conditions fragmentation per se amplifies the negative effect of forest loss (@haddad_habitat_2015; @fletcher_jr_is_2018; @puttker2020indirect; @gonccalves2025species) or, in rare cases, partially buffers it (@fahrig_ecological_2017; @fahrig_SLOSS2020).
</p>


## A mechanistic measure of biodiversity vulnerability

<p>
Within our framework, landscape effects are quantified as changes in the species input parameter (U) needed to approximate the observed local species richness under alternative landscape configurations. Because U quantifies the external species input required to maintain observed local richness (@condit2012; @may_dispersal_2012; @azaele2016), differences in U among experimental landscapes provide a mechanistic measure of how landscape structure modifies the demographic dependence of local communities on external biodiversity input. The species input is interpreted as long-distance dispersal, propagule bank replenishment, and speciation (@ALONSO2006; @etienne2007zero; @jabot2008reconciling).
</p>

<p>
This metric provides a direct link between landscape configuration, functional connectivity, and biodiversity persistence. Landscape effects therefore emerge from contrasts among alternative connectivity regimes rather than from correlations between biodiversity patterns and landscape metrics (@Marshall2014; @frazier2017; @DAGinfEPI2019; @gianluca2022ch4). Because these contrasts are estimated for individual plots, the framework captures heterogeneity in landscape responses that would otherwise be averaged across plots (@Feller2015).
</p>

## Forest loss dominates, but fragmentation shapes variability in the total forest loss and fragmentation effect

<p>
Forest cover loss was consistently associated with increased vulnerability in low-cover landscapes, whereas its effects were negligible in landscapes with high remaining forest cover (@swift2010critical; @villard_beyond_2014; @rivaEbanskleite2024; @haddad_habitat_2015; @fletcher_jr_is_2018; @puttker2020indirect; @gonccalves2025species; @fahrig_ecological_2017; @fahrig_SLOSS2020). The magnitude of this effect was lowest under intermediate dispersal limitation, mirroring the non-linear relationship between dispersal and U observed in pristine landscapes (@may_dispersal_2012). Under strong dispersal limitation, recruitment is dominated by nearby conspecifics, reducing stochastic species loss and therefore the need for external replenishment. As dispersal increases, heterospecific replacement becomes more frequent, increasing drift-driven species loss and raising U. Under weak dispersal limitation, however, distant populations increasingly contribute propagules to local communities, reducing local extinction and causing U to decline, at the cost of reducing landscape total biodiversity (@Mouquet2003Community). This mechanism may also explain why vulnerability to habitat loss was minimized under intermediate dispersal limitation. Because U is highest under these conditions, local diversity becomes less dependent on surrounding habitat as a source of biodiversity replenishment, weakening the ecological consequences of landscape structure (@campos2012, campos2013).
</p>

<p>
In contrast to forest cover loss, fragmentation per se exhibited weak and highly variable effects across landscapes (@fahrig_ecological_2017; @fahrig2019habitatA; @fahrig_SLOSS2020). Fragmentation __per se__ (FPS) effects were not organized by a consistent interaction between dispersal limitation and forest cover, unlike the pattern observed for habitat loss. Positive and negative FPS effects occurred across the forest-cover gradient, with a few plots showing positive responses under strong dispersal limitation and others exhibiting strongly negative responses at intermediate to high forest cover. This pattern is broadly consistent with the long-standing inconsistency of fragmentation responses reported across empirical systems, whereas fragmentation __per se__ has been associated with positive, neutral, and negative responses depending on the study (@fahrig_ecological_2017; @fahrig_SLOSS2020). Our results suggest that this variability emerges because fragmentation effects depend on how particular landscape configurations modify effective connectivity under different dispersal constraints. Under severe dispersal limitation, some spatial configurations may favor local conspecific replacement, reducing stochastic species loss and potentially generating positive fragmentation effects (@campos2012; @campos2013). Conversely, other configurations may further restrict propagule exchange among local communities, amplifying biodiversity loss (@pardini_beyond_2010; @villard_beyond_2014; @campos2012; @campos2013). The scale-sensitivity analysis further indicates that these connectivity dynamics depend not only on dispersal limitation and landscape configuration but also on local community properties (Appendix SI2). Species richness and individual density influence the baseline value of U required to maintain local diversity and therefore affect how landscape configuration alters community dynamics. Consequently, fragmentation effects may emerge from the interplay among spatial configuration, dispersal processes, and local demographic conditions, helping explain why responses remain difficult to generalize across landscapes.
</p>

<p>
The total effect of forest loss and fragmentation (FLF) mainly reflected the patterns observed for forest cover loss in highly deforested landscapes. Consistent with this result, FLF was best explained by the interaction between dispersal limitation and forest cover, indicating that habitat loss remained the primary determinant of vulnerability across landscapes (@swift2010critical; @villard_beyond_2014; @rivaEbanskleite2024; @haddad_habitat_2015; @fletcher_jr_is_2018; @puttker2020indirect; @gonccalves2025species; @fahrig_ecological_2017; @fahrig_SLOSS2020). However, FLF also inherited the variability associated with fragmentation _per se_. In most low-cover landscapes, negative FPS effects amplified the vulnerability associated with habitat loss, producing stronger total effects (@haddad_habitat_2015; @fletcher_jr_is_2018; @puttker2020indirect; @gonccalves2025species). In a smaller number of cases, positive FPS effects partially compensated for habitat-loss vulnerability, particularly under severe dispersal limitation (@fahrig_ecological_2017; @fahrig_SLOSS2020). Together, these results suggest that habitat loss primarily determines the direction of the total landscape effect, whereas fragmentation per se contributes to its variability (@swift2010critical; @villard_beyond_2014; @rivaEbanskleite2024; @haddad_habitat_2015; @fletcher_jr_is_2018; @puttker2020indirect; @gonccalves2025species; @fahrig_ecological_2017; @fahrig_SLOSS2020).
</p>

## Implications for the fragmentation debate

<p>
Our results revealed a general tendency for forest cover loss to dominate landscape effects in highly deforested landscapes, with fragmentation __per se__ contributing additional variability across the gradient of habitat loss. However, the most notable pattern was the variability among plots. Consistent with this result, statistical models that ignored plot-level variation in dispersal responses showed lower explanatory power. In this context, explaining variation among communities may be as important as estimating average landscape effects across a biome.
</p>


<p>
This distinction reflects a broader tension within landscape ecology regarding how biodiversity responses should be incorporated into conservation planning. One tradition emphasizes generalizable heuristics derived from broad comparative analyses of landscape structure and biodiversity responses, prioritizing scalable recommendations based on relatively simple landscape descriptors (e.g., @fahrig_rethinking_2013; @watling2020support; @arroyo-rodriguez_designing_2020; @riva2024overcoming). In contrast, more mechanistic approaches seek to represent species responses, dispersal dynamics, and restoration scenarios explicitly through spatially detailed models tailored to particular ecological contexts (@banks-leite_suggestion_2021; @fletcher2026resurrecting). More broadly, these traditions reflect different strategies for balancing ecological realism, inferential generality, and data availability in conservation science.
</p>


<p>
The SENM framework provides a possible integration between these approaches by explicitly incorporating dispersal limitation and landscape configuration while remaining parameterizable from biodiversity and landscape data that are broadly available across fragmented systems. It allows alternative connectivity scenarios to be evaluated explicitly through counterfactual landscape simulations and empirical evaluation.
</p>

## Limitations

### Orthogonality and decomposition of landscape effects

<p>
Within our framework, the clumped scenario allows the effects of spatial fragmentation to be evaluated independently from overall forest cover (@fahrig_effects_2003; @fahrig_resolving_2022). This formulation also clarifies an inferential problem underlying the fragmentation debate. Once habitat is spatially distributed across a landscape, habitat removal inevitably alters the spatial configuration of the remaining habitat (@didham_rethinking_2012; @fletcher_jr_is_2018; @fahrig2019habitatA; @vancine2024atlantic; @fletcher2026resurrecting). In this sense, the fragmentation debate may be interpreted less as a disagreement about whether fragmentation matters and more as a question about the conditions under which forest loss and fragmentation effect can be meaningfully decomposed into independent ecological components (@fahrig_effects_2003; @didham_rethinking_2012; @fletcher2018habitat; @fahrig2019habitatA; @puttker2020indirect; @watling2020support). Complete orthogonality between habitat amount and fragmentation therefore represents a limiting inferential case rather than an intrinsic property of empirical landscapes (@pearl2016causal).
</p>


<p>
Decomposing the effects of forest cover loss and fragmentation _per se_ also requires specifying the spatial scale over which this decomposition is defined (@jackson_are_2015; @miguet2016determines; @fletcher_jr_addressing_2023). In our framework, this scale is determined by the spatial extent from which habitat is redistributed to construct the clumped landscape. By contrast, estimating the combined effect of forest loss and fragmentation only requires a landscape extent sufficiently large to avoid constraining dispersal processes, because SENM simulations are not inherently limited by landscape boundaries (@thompsonPycoalescence2020). Consequently, decomposing landscape effects requires stronger assumptions than estimating their combined effect. These assumptions become unavoidable when information about ecologically relevant spatial scales is unavailable (@jackson_are_2015; @miguet2016determines; @fletcher_jr_addressing_2023). In such cases, the inferred contributions of forest cover loss and fragmentation per se depend increasingly on how contemporary biodiversity patterns are assumed to relate to landscape structure (@frazier2017).
</p>


### Temporal correspondence between biodiversity and landscape patterns

<p>
Most observational studies relate contemporary communities to contemporary landscape configuration (@frazier2017), implicitly assuming that the landscape conditions shaping community assembly have remained sufficiently stable through time or that delayed ecological responses have largely relaxed (@kuussaari_extinction_2009; @lira_temporal_2019; @figueiredo_understanding_2019). This assumption is common to landscape analyses based on biodiversity and landscape snapshots because historical information on habitat dynamics is often far more limited than the timescales over which ecological responses to habitat change unfold (@de_lima_how_2015; @de2020erosion; @mapbiomas; @chase2019fragsad).
</p>
 
<p>
Within the SENM framework, this assumption is operationalized through ecological equilibrium (@ALONSO2006; @rosindell2012case; @azaele2016). Under ecological equilibrium, biodiversity patterns are assumed to reflect the long-term interaction between dispersal processes and landscape structure rather than transient responses to recent landscape change. Consequently, violations of this assumption may contribute to deviations between observed and simulated communities. In this context, SENM simulations under pristine and fragmented landscapes can be interpreted as reference scenarios in which the observed community is assumed to be in equilibrium with the pre-loss landscape (no extinction debt paid) or with the contemporary landscape (extinction debt fully paid), respectively (@thompson2019characterising). The clumped landscape provides a third reference in which forest area is lost but isolation is minimized, approximating the expectation that local diversity depends on habitat amount rather than arrangement (@etienne2005new; @fahrig_rethinking_2013).
</p>

<p>
Nevertheless, contemporary landscape configuration may retain information about the historical processes that generated present-day habitat patterns. Deforestation trajectories are not spatially random: road expansion, agricultural colonization, and successive cycles of land use and forest regeneration produce characteristic spatial signatures that persist in contemporary landscapes, from the fishbone deforestation pattern of the Amazon to the heterogeneous mosaic created by centuries of disturbance and regeneration in the Atlantic Forest (@de2006thresholds; @lira2012land; @Wagner2020Mapping; @Jakovac2021The; @alencar2023forest; @Broggio2024Quantifying). Consequently, contemporary landscape structure may encode part of the historical context relevant to community assembly even when explicit temporal data are unavailable.
</p>

<p>
Consistent with this interpretation, variation in model congruence among plots was largely explained by the interaction between dispersal limitation and experimental landscape type at the plot level. Moreover, accounting for this interaction eliminated residual spatial autocorrelation among plots, suggesting that contemporary landscape structure and dispersal jointly capture meaningful spatial variation across the Atlantic Forest. This suggests that the simulated connectivity process was sufficient to explain spatial variation in congruence without requiring explicit geographic coordinates. Although these findings do not directly test historical landscape dynamics (@fletcher2026resurrecting), they support the idea that present-day landscapes retain some information relevant to mechanistic inference about biodiversity patterns.
</p>

## Future directions

<p>
Rather than defending neutrality as a description of tropical tree communities, we see the SENM as a first, tractable approximation, to be refined where and when is possible (@rosindell2012case; @leroi2020). Earlier assessments led to the expectation that neutral models would rarely reproduce biodiversity patterns in fragmented forests (@Gilbert2006; @laurance2008theory; @etienne_spatial_2011). Yet most of our plots showed good SAD congruence under at least one combination of dispersal limitation and landscape, although static patterns cannot address the temporal turnover examined by @Gilbert2006. Because richness and SADs carry little information about competitive hierarchies, this result does not indicate that species are equivalent (@adler2007niche; @chisholm_niche_2010; @hammal_when_2015). It indicates that these patterns cannot, by themselves, discriminate among such hypotheses about species differences, but can elucidate the influence of landscape features in the overall tree demography (@campos2012; campos2013; claudino_extinction_2015; @chisholm_speciesarea_2018; @thompson2019characterising; the present study). The question then shifts from whether neutrality holds to which auxiliary assumptions matter most for the inferences we want to draw, and how to relax them without losing practical value (@rosindell2012case; @bausman_aims_2019; @leroi2020; @thompsonPycoalescence2020). In this approach, plots where the SENM was systematically incongruent are not simply discarded; they are where refinements can be tested.
</p>

<p>
For landscape effects, the most consequential auxiliary assumption is the matrix (@fletcher_jr_is_2018; @fahrig2019habitatA; @puttker2020indirect; @fahrig_resolving_2022; @vancine2024atlantic), which is currently inert: demography does not change near edges, and dispersal through the matrix equals dispersal through habitat. Relaxing it follows the logic applied here to fragmentation. Because habitat is spatially arranged, its loss necessarily alters configuration, so the decomposition of loss and fragmentation __per se__ is defined only relative to an explicit reference landscape (@didham_rethinking_2012; @fahrig_resolving_2022). The matrix adds a further layer. In a binary map, matrix amount is the complement of habitat amount and cannot vary independently of it, but the functional properties of the converted land can, notably edge effects on mortality and recruitment and resistance to dispersal (@laurance2008theory; @bello2015defaunation; @melito2018landscape; @schwartz2019fragmentation; @willmer2022global). These properties are also products of the conversion process. Land is converted deliberately to exploit properties of the converted area or to gain access to remaining habitat, and habitat loss, fragmentation and matrix effects are consequences of that process, each acting through different demographic pathways (@didham_rethinking_2012; @fahrig_resolving_2022; @magliocca2023review; @ekstrom2024modelling; @fletcher2026resurrecting).
</p>

<p>
This suggests a design with six types of experimental landscapes (Fig. 3). Two landscapes add a single matrix effect, edge or dispersal friction, and a third combines both, forming a 2 × 2 crossed design within the fragmented configuration. Contrasts with the fragmented give the effects of edge, friction and their interaction, and contrasts between the pristine and each matrix landscape give the total effect of habitat loss and matrix conversion, equal to FLF plus the matrix effect. Edge effects are caused by changes in tree demography near forest edges (@laurance2008theory; @melito2018landscape; @schwartz2019fragmentation), and friction in reduced seed dispersal through degraded matrices (@bello2015defaunation; @emer2020seed; @hatfield2024constraints). Neither matrix effects are binary: friction can be varied as dispersal limitation was, and edge effects require an influence function describing how mortality changes with distance to the matrix, in both magnitude and depth. These parameters could be constrained by empirical syntheses and treated as scenarios.
</p>

<p>
Other auxiliary assumptions rank lower in this sequence because relaxing them is costlier. To relax equilibrium, zero-sum and the static landscape configuration assumptions (@etienne2007zero; @aguilar2019habitat; @martensen2017) there will be necessary to explore alternative algorithms to the coalescent approach of the SENM and probability will increase considerably the running time of each individual simulation (@rosindell2008;  @thompsonPycoalescence2020). They would multiply parameters and computation, and would require information on landscape and community history (@metzger2017best; @pereira2020developing; @fletcher2026resurrecting). 
</p>

# Conclusion (254)

<p>
Landscape effects on tree communities may be inferred through counterfactuals. We isolated forest loss and fragmentation per se by contrasting experimentally constructed landscapes, and grounded the interpretation of these contrasts in an explicit demographic mechanism: in the SENM, dispersal across landscape configuration generates connectivity and its consequences for local biodiversity. Calibration against observed richness and validation against species abundance distributions delimited the scope of this inference, while retaining the majority of plots, which did not differ environmentally from those excluded.
</p>

<p>
Variation among plots was a central feature of how landscape structure affected biodiversity. Within the dispersal range retained for interpretation, forest cover loss dominated the total vulnerability effect in low-cover landscapes, with the strongest effects at both ends of the dispersal range. Fragmentation __per se__ was generally weaker and more variable, and our results suggest that it may depend on how a given configuration interacts with dispersal limitation and local community properties (e.g. species richness and individual density), rather than on configuration alone.
</p>

<p>
These estimates support a situated, mechanism-based counterfactual interpretation. This interpretation is conditional on the SENM assumptions, the dispersal scenarios, and the empirical evidence used to calibrate and evaluate the model. These estimates also remain conditional on an inert matrix and equilibrium, both explicit targets for future refinement. By making the counterfactual contrasts and their underlying demographic mechanisms explicit, the framework treats variation among plots as part of the landscape effect to be explained rather than averaged out. This provides a way to connect general conservation heuristics with context-specific mechanistic evidence.
</p>


# References


```{=latex}
\bibliographystyle{plainnat}
\bibliography{resources/citations2}
```


<!-- código para construção e display dos eleentos textuais -->

# Table

```{r caption tab-selecao,echo=FALSE,include=TRUE,eval=TRUE}
library(officer)
f_z <- function(x) (x-mean(x))/sd(x)
f_label <- \(vn){
  gsub("1","M4",vn) %>% 
    gsub("s\\(k,by\\=land\\) \\+ \\(lat,long\\)","M1",.) %>% 
    gsub("s\\(k,by\\=land\\) - te\\(k,plot,by\\=land\\)","M3",.) %>% 
    gsub("s\\(k,by\\=land\\)","M2",.)
    
}
f_antilabel <- \(vn){
  gsub("M4","1",vn) %>% 
    gsub("M1","s\\(k,by\\=land\\) \\+ \\(lat,long\\)",.) %>% 
    gsub("M3","s\\(k,by\\=land\\) - te\\(k,plot,by\\=land\\)",.) %>% 
    gsub("M2","s\\(k,by\\=land\\)",.)
    
}
vcap <- "Model comparison among candidate hierarchical generalized additive models (HGAMs). ΔAICc represents the difference in AICc relative to the best-supported model; est. coef. denotes the effective degrees of freedom (model complexity); weight indicates Akaike weight; Dev. Exp. represents deviance explained. Model formulas are expressed using the mgcv package syntax: s() denotes a univariate smooth term, and te() denotes a tensor-product smooth for interactions between continuous variables (here, dispersal limitation, k, and landscape forest cover, p). The notation | Plot indicates plot-specific smooth terms (hierarchical structure), allowing dispersal effects to vary among plots."
```
```{r tab-selecaoCong,echo=FALSE,include=TRUE,eval=TRUE}
df_tabsel <- readRDS(file="1_to_compile_dissertacao_EM_USO/09_SI/RDS/tabsel_cong_logUU.rds") %>% 
  mutate(dev.expl=100*dev.expl,
         across(where(is.numeric), ~round(.x,2)),
         Effect=base,
         modelo=gsub("Site","Plot",modelo)) %>% 
  select(-base)
library(flextable)
f_label <- \(vn){
  gsub("1","M4",vn) %>% 
    gsub("s\\(k,by\\=land\\) \\+ \\(lat,long\\)","M1",.) %>% 
    gsub("s\\(k,by\\=land\\) - te\\(k,plot,by\\=land\\)","M3",.) %>% 
    gsub("s\\(k,by\\=land\\)","M2",.)
    
}
f_flextable <- \(dfi){
  #
  df_grouped <- as_grouped_data(dfi, groups = "Effect")
  #
  as_flextable(df_grouped) %>%
    # Estética geral
    bg(bg = "white", part = "all") %>%
    theme_vanilla() %>% 
    # 
    set_header_labels(
      modelo = "HGAM",
      dAICc = "ΔAICc",
      df = "est. coef.", # 
      weight = "weight",
      dev.expl = "Dev. Exp."
    ) %>%
    
    # Destacar as linhas que são os nomes dos grupos (Frag. total, etc)
    bold(i = ~ !is.na(Effect), bold = TRUE, part = "body") %>%
    #as_italic(i = ~ !is.na(base), part = "body") %>%
    bg(i = ~ !is.na(Effect), bg = "#F7F7F7", part = "body") %>% # 
    
    # Formatação de números
    colformat_double(digits = 2) %>% 
    colformat_double(j = "weight", digits = 2) %>% # 
    
    # Alinhamento e legenda
    align(align = "center", part = "all") %>% 
    align(j = 1, align = "left", part = "body") %>% # 
    set_caption(caption = vcap) %>% 
    # add_header_lines(values = as_paragraph(
    #         as_chunk("Table 1. ", props = fp_text(bold = TRUE)),
    #         as_chunk(vcap, 
    #                  props = fp_text(bold = FALSE),
    #                  fp_p = fp_par(text.align = "justify") 
    #                  )
    #          )
    #         ) %>%
    # 
    set_table_properties(layout = "fixed") %>%
    # Define a largura apenas para a coluna 'modelo'
    flextable::width(j = "modelo", 
                     width = 10.5/5, unit = "in") %>%
    # Define a largura para todas as OUTRAS colunas
    flextable::width(j = c("dAICc", "df", "weight", "dev.expl"), 
                     width = 1, unit = "in")
}
fxtow <- f_flextable(df_tabsel)
# saveRDS(fxtow,file = "1_to_compile_dissertacao_EM_USO/09_SI/RDS/fxtow_tab1.rds")
# save_as_image(fxtow, 
#               path = "figuras/tabsel_efeitos.png")
fxtow
```


# Figures

```{r criação da figCong, eval=FALSE,include=FALSE,echo=FALSE}
l_dfpred <- readRDS(file = "dados/csv_SoE/rds/l_dfpred_md_cong_absoluta.rds")
library(hexbin)
p <- l_dfpred$fixo_e_aleat %>% 
  ggplot(aes(x=k,y=nSAD)) +
  # geom_point(alpha=0.3) +
  geom_line(aes(y=fit,group = SiteCode),alpha=0.3) +
  # geom_boxplot(aes(group=k),alpha=0.6) +
  geom_ribbon(data=l_dfpred$fixo,
              aes(x=k,y=fit,ymin=lower,ymax=upper),
              color="blue",
              fill="lightblue") +
  geom_line(data=l_dfpred$fixo,
            aes(x=k,y=fit),
            color="darkred") +
  geom_hex(bins = 50,alpha=0.5) + 
  scale_fill_gradient("count",low = "gray", high = "black", na.value = NA) +
  scale_x_continuous(expand = c(0.01, 0)) +
  scale_y_continuous(expand = c(0.01, 0)) +
  labs(x="Fraction of propagules within the immediate neighborhood (k)",
       y="Number of congruent simulated SADs (goodness-of-fit)" #,
       # title="Congruence of simulated SADs across experimental landscapes",
       # subtitle="Congruence is best described by connectivity and spatial proximity among forest plots"
       ) +
  facet_wrap(~land) +
  theme_bw() +
  theme(strip.text = element_text(size=12, #margin=margin(),
                                  face="bold"))
saveRDS(p,file = "1_to_compile_dissertacao_EM_USO/09_SI/RDS/p_interpretacao_cong.rds")
ggsave(filename="figuras/descricao_congruencia_absoluta.png",
       p,height = 5,width = 13.8)
library(magick)
img <- image_read("figuras/descricao_congruencia_absoluta.png") %>% 
  image_resize("50%") %>% 
  image_trim()
image_write(img,path = "figuras/descricao_congruencia_absoluta.png")
```
```{r captio figCong,eval=TRUE,echo=FALSE}
vcap <- "Congruence between simulated and observed SADs (y-axis) across dispersal (x-axis) and landscape combinations (facet labels). Hexagonal bins show the density of plot‑level observations (lighter = fewer points, darker = up to 100 points). Gray lines: plot‑level predictions from the most plausible model with spatial coordinates set to zero. Red line: overall mean prediction; blue shading: 95% pointwise confidence interval. Mean congruence exceeded 75% in most simulations but declined under very weak dispersal limitation, especially in clumped and pristine landscapes."
```
```{r figCong,fig.cap=vcap,fig.width=14,fig.height=5.5,eval=TRUE,include=TRUE}
p <- readRDS(file = "1_to_compile_dissertacao_EM_USO/09_SI/RDS/p_interpretacao_cong.rds")
p
```

```{r criacao de figura pefeitos,eval=FALSE,include=FALSE,echo=FALSE}
library(metR)
library(gridExtra)
library(ggplot2)
library(dplyr)
#-------------------------------------1----------------------
#--------------------- base de dados  ----------------------
#-----------------------------------------------------------
df_real <- readRDS("dados/csv_SoE/rds/df_obs_pred_plot_logUU_pk.rds") %>% 
  mutate(
    p_class = case_when(
       p>=0.99 ~ "%FC ≥ 99",
       p<0.99 & p>=0.80 ~ "80 ≤ %FC < 99",
       p<0.80 & p>=0.60 ~ "60 ≤ %FC < 80",
       p<0.60 & p>=0.30 ~ "30 ≤ %FC < 60",
       p<0.30 ~ "%FC < 30"),
    p_class = factor(p_class,levels=c("%FC < 30",
                                           "30 ≤ %FC < 60",
                                           "60 ≤ %FC < 80",
                                           "80 ≤ %FC < 99",
                                           "%FC ≥ 99")),
    lower = fit - 1.96 * se.fit,
    upper = fit + 1.96 * se.fit
  )
#-----------------------------------------------------------
#------------------ função da figura -----------------------
#-----------------------------------------------------------
f_obs_predito_bysite <- \(dff,logic_ribbon=FALSE,vpos=c(0.72,0.25),vt=15,vts=0.90){
  dfpred <- dff %>% 
    mutate(
      p_class = case_when(
        p>=0.99 ~ "%FC ≥ 99",
        p<0.99 & p>=0.80 ~ "80 ≤ %FC < 99",
        p<0.80 & p>=0.60 ~ "60 ≤ %FC < 80",
        p<0.60 & p>=0.30 ~ "30 ≤ %FC < 60",
        p<0.30 ~ "%FC < 30"),
      p_class = factor(p_class,levels=c("%FC < 30",
                                        "30 ≤ %FC < 60",
                                        "60 ≤ %FC < 80",
                                        "80 ≤ %FC < 99",
                                        "%FC ≥ 99")),
      lower = fit - 1.96 * se.fit,
      upper = fit + 1.96 * se.fit
    )
  df_ij <- dfpred %>% 
    select(SiteCode,p_class) %>% 
    distinct() %>% 
    group_by(p_class) %>% 
    tally() %>% 
    mutate(label = paste0(p_class,", n = ",n)) %>% 
    select(-n)
  df_ij$label <- factor(df_ij$label,
                        levels=unique(df_ij$label))
  dfp <- inner_join(dfpred,df_ij)
  nsites <- filter(dfp,p_class=="%FC ≥ 99") %>% pull(SiteCode) %>% unique %>% length()
  dfp <- mutate(dfp,
                efeito = gsub("FLF","Forest Loss and Frag. (FLF)",efeito) %>% 
                  gsub("FPS","Frag. per se (FPS)",.) %>% 
                  gsub("LFC","Loss of Forest Cover (LFC)",.)
                )
  dfrange <- filter(dfp,p_class=="%FC ≥ 99") %>% 
    reframe(
      names = c("amplitude","amplitude","Q 5%-95%","Q 5%-95%"),
      value = c(min(Uefeito), max(Uefeito),quantile(Uefeito,probs=0.05),quantile(Uefeito,probs=0.95))
    )
  dfp <- filter(dfp,p_class!="%FC ≥ 99")
  p <- dfp %>% 
    mutate(efeito = factor(efeito,
                           levels=c("Forest Loss and Frag. (FLF)",
                                    "Frag. per se (FPS)",
                                    "Loss of Forest Cover (LFC)"))) %>% 
    ggplot(aes(x=k,y=Uefeito)) +
    geom_boxplot(aes(group=k),alpha=0.6) +
    geom_hline(yintercept = 0,color="black") +
    geom_hline(data = dfrange, aes(yintercept = value,linetype=names),
               color="black",linewidth=0.5) +
    scale_linetype_manual(name = paste0("%FC ≥ 99",", n = ",nsites),
                          values=c("dotted","dashed")) +
    {if(logic_ribbon)geom_ribbon(aes(x=k,y=fit,ymin=lower,ymax=upper,group = SiteCode),
                                 fill="lightblue",
                                 alpha=0.2)}+
    geom_line(aes(y=fit,group = SiteCode,color=p),alpha=0.4) + #,color="darkred"
    geom_point(aes(color=p),alpha=0.85) +
    scale_colour_gradient2("% FC",midpoint=0.5,
                           low="red",
                           mid = "yellow",
                           high = "darkgreen") +
    labs(x="Fraction of propagules dispersing locally (k)",y="logU/U",title="") +
    facet_grid(efeito~label,scales="free") +
    theme_linedraw() +
    theme(plot.margin=unit(c(0,0.2,0,0), "cm"),
          text=element_text(size=vt,face="bold"),
          strip.background = element_rect(fill = "white"),
          strip.text = element_text(color="black"),
          legend.position = "inside",
          legend.position.inside = vpos,
          legend.direction="horizontal",
          legend.box = "horizontal",
          legend.box.just = "left",
          legend.background = element_rect(color="black")
          ) +
    guides(linetype = guide_legend(
      title.theme = element_text(size = vt*vts, face = "bold")
      )
    )
  return(list("plot"=p,"range"=dfrange))
}
#
#-----------------------------------------------------------
#------------------ criação da figura ----------------------
#-----------------------------------------------------------
lret <- f_obs_predito_bysite(df_real,logic_ribbon = TRUE, vpos=c(0.72,0.05),vt = 12.5)
p_efeitos <- lret$plot
vfac <- 1.2
ggsave(filename = "1_to_compile_dissertacao_EM_USO/00_Resultados/figuras/p_efeitos_porclasseCF.png",
       dpi=200,
       plot=p_efeitos,
       width=12*vfac,height=9*vfac)
saveRDS(p_efeitos,file="1_to_compile_dissertacao_EM_USO/00_Resultados/figuras/p_efeitos_porclasseCF.rds")
dfrange <- lret$range
saveRDS(dfrange,file="1_to_compile_dissertacao_EM_USO/00_Resultados/figuras/dfrange.rds")
```
```{r vcap p-efeitos,eval=TRUE,echo=FALSE}
dfrange <- readRDS(file="1_to_compile_dissertacao_EM_USO/00_Resultados/figuras/dfrange.rds")
vlabel <- sapply(dfrange$value,round,digits=3)
names(vlabel) <- paste0(rep(c("neg","pos"),2),dfrange$names)
vcap <- glue("Proposed landscape effects described by the functional metric log Ui/Uj, shown as observed values and predictions from a hierarchical generalized additive model (HGAM). Differences in U reflect variation in the external replenishment required to sustain observed diversity patterns. The metric log(Ui/Uj) quantifies the relative change in this requirement and can thus be interpreted as a measure of landscape-induced vulnerability. Each point represents the comparison between two simulations, varying according to the landscape effect under evaluation (row titles). Forest Loss and Fragmentation (FLF): pristine vs  fragmented (joint change in forest amount and configuration); Fragmentation per se (FPS): clumped vs fragmented (change in configuration, constant forest amount); Loss of Forest Cover (LFC): pristine vs clumped (change in forest amount, constant aggregation). The x-axis represents dispersal limitation (k), the fraction of propagules dispersed within the immediate neighbourhood of the parent individual. Points are coloured according to the percentage of forest cover (%FC) observed within the 2-km landscape surrounding each focal plot. Columns correspond to classes of observed %FC for each plot (column titles). Plots with greater and equal 99% forest cover were excluded from the graphical display; only the amplitude (maximum = 0.075; minimum = -0.074) and the central 90% interval (95th quantile = 0.04; 5th quantile = -0.038 ) of log Ui/Uj are indicated on the y-axis. This variability reflects the stochasticity inherent to the spatially explicit neutral model. Boxplots summarise values for each level of k. The y-axis scale differs among landscape effects. Only plots classified as congruent are shown.")
```
```{r p-efeitos, fig.cap=vcap,fig.width=14*1.2,fig.height=10*1.2}
p_efeitos <- readRDS("1_to_compile_dissertacao_EM_USO/00_Resultados/figuras/p_efeitos_porclasseCF.rds")
p_efeitos
```


```{r vcap fig3}
vcap <- "Experimental landscapes used (Pristine, Clumped, Fragmented) and proposed (Matrix friction, Matrix edge, Matrix combined) to estimate effects of forest loss, fragmentation and matrix conversion. Each landscape is a schematic 8 × 8 mosaic of habitat (green) and matrix (grey), not to scale. Clumped, Fragmented and the three matrix landscapes have the same amount of habitat; Pristine has none lost. In the matrix landscapes, edge effects alter demography in habitat cells adjacent to the matrix (orange), and friction reduces dispersal through matrix cells (dark grey); Matrix combined has both. Arrows run from the landscape in the numerator to the landscape in the denominator of each contrast, log(U_numerator/U_denominator): forest loss and fragmentation (FLF = log[U_pristine/U_fragmented]), fragmentation per se (FPS = log[U_clumped/U_fragmented]), forest cover loss (LFC = log[U_pristine/U_clumped]) and matrix effects (M = log[U_fragmented/U_matrix landscape]). Because contrasts are log-ratios, they are additive: FLF = FPS + LFC, and the total effect of habitat conversion to matrix is FLF + M; the interaction between edge and friction is M_combined − M_friction − M_edge. Friction and edge are shown at a single intensity, but both may vary continuously."
```
```{r fig3display, eval=TRUE,include=TRUE,echo=FALSE,fig.cap=vcap,fig.width=6.8,fig.height=6.8}
# Esquema das paisagens experimentais (grade 3 x 3)
# Posição (linha x coluna): Pristine (1x1), Clumped (1x2), Fragmented (2x2),
#                           Mdisp (2x3), Medge (3x2), Mde (3x3)
# Autor do esquema: Danilo P. Mori | Script: ggplot2 apenas (sem dplyr)
# setwd("~/Documentos/artigo_principal/")
library(ggplot2)

step <- 18   # distância entre origens das células da grade
n    <- 8    # cada paisagem é um mosaico n x n (apenas esquemático)

## 1. Mapas binários (1 = habitat, 0 = matriz) ---------------------------
pristine <- matrix(1L, n, n)

frag_chr <- c("11100000",
              "11100110",
              "00000110",
              "01100000",
              "01100011",
              "00000011",
              "10110000",
              "10110000")
fragmented <- do.call(rbind, lapply(strsplit(frag_chr, ""), as.integer))

clumped <- matrix(0L, n, n)
clumped[3:6, 2:7] <- 1L          # bloco central 4 x 6 = 24 células

# mesma quantidade de habitat em Clumped e Fragmented (condição do desenho)
stopifnot(sum(fragmented) == sum(clumped))

## 2. Borda: habitat com ao menos 1 vizinho (4-conectado) de matriz ------
is_edge <- function(m) {
  nr <- nrow(m); nc <- ncol(m)
  e <- matrix(FALSE, nr, nc)
  for (i in seq_len(nr)) for (j in seq_len(nc)) {
    if (m[i, j] == 1L) {
      nb <- list(c(i - 1, j), c(i + 1, j), c(i, j - 1), c(i, j + 1))
      e[i, j] <- any(vapply(nb, function(p) {
        p[1] >= 1 && p[1] <= nr && p[2] >= 1 && p[2] <= nc && m[p[1], p[2]] == 0L
      }, logical(1)))
    }
  }
  e
}

## 3. Tabela de células de uma paisagem ------------------------------------
lev <- c("Habitat", "Habitat (edge effect)", "Matrix", "Matrix (friction)")

glyph_df <- function(m, id, grow, gcol, edge = FALSE, friction = FALSE) {
  e  <- if (edge) is_edge(m) else matrix(FALSE, nrow(m), ncol(m))
  ij <- expand.grid(i = seq_len(nrow(m)), j = seq_len(ncol(m)))
  hab <- m[cbind(ij$i, ij$j)] == 1L
  edg <- e[cbind(ij$i, ij$j)]
  ij$type <- ifelse(hab & edg, lev[2],
             ifelse(hab,       lev[1],
             ifelse(friction,  lev[4], lev[3])))
  ij$x  <- (gcol - 1) * step + ij$j
  ij$y  <- -(grow - 1) * step - ij$i
  ij$id <- id
  ij
}

lands <- rbind(
  glyph_df(pristine,   "Pristine",   1, 1),
  glyph_df(clumped,    "Clumped",    1, 2),
  glyph_df(fragmented, "Fragmented", 2, 2),
  glyph_df(fragmented, "Mdisp",      2, 3, friction = TRUE),
  glyph_df(fragmented, "Medge",      3, 2, edge = TRUE),
  glyph_df(fragmented, "Mde",        3, 3, edge = TRUE, friction = TRUE)
)
lands$type <- factor(lands$type, levels = lev)

## 4. Molduras e nomes -----------------------------------------------------
panels <- data.frame(
  id   = c("Pristine", "Clumped", "Fragmented", "Mdisp", "Medge", "Mde"),
  grow = c(1, 1, 2, 2, 3, 3),
  gcol = c(1, 2, 2, 3, 2, 3),
  name = c("Pristine", "Clumped", "Fragmented",
           "Matrix friction", "Matrix edge", "Matrix combined")
)
panels$ox <- (panels$gcol - 1) * step
panels$oy <- -(panels$grow - 1) * step
# panels$base <- panels$id == "Fragmented"      # paisagem basal em moldura grossa

## 5. Setas: sentido = direção da alteração da paisagem -------------------
arr <- data.frame(
  x    = c( 9.2, 22.5, 9.4,  27.2, 22.5, 40.5,  27.2, 27.2),
  y    = c(-4.5, -11.1, -9.4, -22.5, -29.1, -29.1, -40.5, -27.2),
  xend = c(16.8, 22.5, 17.6,  34.8, 22.5, 40.5,  34.8, 35.0),
  yend = c(-4.5, -17.8, -17.6, -22.5, -35.8, -35.8, -40.5, -35.0)
)
arr_lab <- data.frame(
  x = c(13.0, 23.3, 12.3, 31.0, 23.3, 41.3, 31.0, 31.6),
  y = c(-3.2, -14.4, -14.8, -21.2, -32.4, -32.4, -39.2, -29.6),
  lab = c("LFC", "FPS", "FLF", "friction", "edge", "+ edge", "+ friction", "both"),
  hj  = c(0.5, 0, 1, 0.5, 0, 0, 0.5, 0)
)

## 6. Legenda e chave de contrastes (células vazias da grade) -------------
leg <- data.frame(x = 37, y = -c(2, 4, 6, 8), type = factor(lev, levels = lev))
key <- paste(
  "Contrasts, log(U numerator / U denominator):",
  "FLF = log(U_pristine / U_fragmented)",
  "FPS = log(U_clumped / U_fragmented)",
  "LFC = log(U_pristine / U_clumped)",
  "M   = log(U_fragmented / U_matrix landscape)",
  "Total (loss + matrix) = FLF + M",
  "Arrows: direction of landscape contrast.",
  "Bold frame: baseline (inert matrix).",
  sep = "\n")
key_df <- data.frame(
  y = -20 - 1.15 * (0:6),
  lab = c(
    "'Contrasts: '*log(U[numerator]/U[denominator])",
    "FLF == log(U[pristine]/U[fragmented])",
    "FPS == log(U[clumped]/U[fragmented])",
    "LFC == log(U[pristine]/U[clumped])",
    "M == log(U[fragmented]/U[matrix~landscape])",
    "'Total (loss + matrix)' == FLF + M",
    "'Arrows: from numerator to denominator'" #,
    # "'Bold frame: baseline (inert matrix)'"
  )
)

pal <- c("Habitat" = "#4C9F70", "Habitat (edge effect)" = "#E69F00",
         "Matrix" = "#E6E6E6", "Matrix (friction)" = "#8C8C8C")

## 7. Figura ---------------------------------------------------------------
p <- ggplot() +
  geom_tile(data = lands, aes(x, y, fill = type),
            width = 1, height = 1, colour = "white", linewidth = 0.25) +
  geom_tile(data = leg, aes(x, y, fill = type),
            width = 1.4, height = 1.4, colour = "grey40", linewidth = 0.2) +
  # geom_rect(data = panels,
  #           aes(xmin = ox + 0.5, xmax = ox + n + 0.5,
  #               ymin = oy - n - 0.5, ymax = oy - 0.5, linewidth = base),
  #           fill = NA, colour = "grey15") +
  geom_text(data = panels, aes(x = ox + (n + 1) / 2, y = oy - n - 1.0, label = name),
            size = 3.4, vjust = 1, fontface = "bold") +
  geom_segment(data = arr, aes(x = x, y = y, xend = xend, yend = yend),
               arrow = arrow(length = unit(2.2, "mm"), type = "closed"),
               colour = "grey25", linewidth = 0.5) +
  geom_text(data = arr_lab, aes(x = x, y = y, label = lab, hjust = hj),
            size = 3.2, colour = "grey15") +
  annotate("text", x = 39, y = -c(2, 4, 6, 8), label = lev,
           hjust = 0, size = 3.0) +
  annotate("text", x = 36.6, y = -0.2, label = "Cell type", hjust = 0,
           vjust = 0, size = 3.2, fontface = "bold") +
  # annotate("text", x = 0.5, y = -20, label = key, hjust = 0, vjust = 1,
  #          size = 2.8, lineheight = 1.15) +
  geom_text(data = key_df, aes(x = 0.5, y = y, label = lab),
            parse = TRUE, hjust = 0, vjust = 1, size = 2.8) +
  scale_fill_manual(values = pal, guide = "none") +
  scale_linewidth_manual(values = c("TRUE" = 1.3, "FALSE" = 0.4), guide = "none") +
  coord_fixed(xlim = c(-0.5, 48), ylim = c(-47.5, 1.5), expand = FALSE) +
  theme_void() +
  theme(plot.background = element_rect(fill = "white", colour = NA),
        plot.margin = margin(4, 4, 4, 4))

# ggsave("figuras/fig3.png", p, width = 8, height = 8, dpi = 300, bg = "white")
# library(magick)
# img <- image_trim(
#   image_read("figuras/fig3.png")
# )
# image_write(img,path = "figuras/fig3.png")
# ggsave("fig_paisagens_matriz.pdf", p, width = 8, height = 8)
# saveRDS(p,file = "fig_paisagens_matriz.rds")
p
```

