---
title: ""
author: ""
output:
  bookdown::pdf_document2:
    latex_engine: lualatex
    pandoc_args: [
      "--number-lines",
      "--variable", "geometry=a4paper",
      "--variable", "geometry=twoside",
      "--variable", "geometry=top=3cm",
      "--variable", "geometry=bottom=2cm",
      "--variable", "geometry=inner=3cm",
      "--variable", "geometry=outer=2cm"
      ]
  bookdown::word_document2: default
  pdf_document: default
  html_document:
    df_print: paged
bibliography: resources/citations2.bib
csl: "resources/ecology-letters.csl"
editor_options:
  chunk_output_type: console
header-includes:
  - \usepackage{polyglossia}
  - \setmainlanguage{english}
  - \renewcommand{\contentsname}{Table of Contents}
  - \usepackage{caption}
  - \captionsetup[figure]{name=Figure, font=it}
  - \captionsetup[table]{name=Table, font=it}
  - \usepackage{tcolorbox}
  - \usepackage{hyperref}
  - \hypersetup{colorlinks=true, linkcolor=blue, citecolor=red, urlcolor=blue}
  - \usepackage[brazilian,hyperpageref]{backref}
  - \tcbuselibrary{breakable}
  - \tcbuselibrary{skins}   
  - \usepackage{lmodern}
  - \usepackage{fontspec}
  - \usepackage{threeparttable} 
  - \usepackage{caption}
  - \usepackage{setspace}
  - \usepackage{etoolbox}
  - \setmainfont{Times New Roman}
  - \fontsize{12}{22}
  - \setlength{\parindent}{1.5cm}
  - \onehalfspacing
  - \makeatletter
  - \patchcmd{\@startsection}{\@afterindentfalse}{\@afterindenttrue}{}{}
  - \makeatother
  - \captionsetup[figure]{font=it}
  - \captionsetup[table]{font=it}
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup resultados,include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = FALSE,eval = TRUE,message = FALSE,warning = FALSE,cache=TRUE,
                      fig.pos = "H", fig.show = "hold",
                      dpi = 60)
library(gridExtra)
library(ggplot2)
library(readr)
library(dplyr)
library(plyr)
library(flextable)
library(officer)
library(magick)
```

```{r trackdown compilacao integral, eval=FALSE, include=FALSE}
library(trackdown)
# both overwrite: 
# overwrite the current google drive file with the current Rmd:
upload_file(file="./1_to_compile_dissertacao_EM_USO/compilacao_integral.Rmd",
            gpath="mestrado/artigo_principal/secoes_texto/",
            hide_code=TRUE)
update_file(file="./1_to_compile_dissertacao_EM_USO/compilacao_integral.Rmd",
            gpath = "mestrado/artigo_principal/secoes_texto/",
            hide_code = FALSE)
# overwrite the current Rmd with the current google drive 
download_file(file="compilacao_integral.Rmd",
              gpath = "mestrado/artigo_principal/secoes_texto/")
```

 
# Título

Dispersal constraints shape the sign and magnitude of biodiversity responses to forest loss and fragmentation

Revista pretendida: Ecography
 
# Resumo {-}  

REFAZER


# Abstract {-}

REFAZER

# Introduction

Habitat loss and fragmentation are major drivers of biodiversity change in tropical forests (REF). Yet empirical estimates of the independent effects of fragmentation remain highly inconsistent, ranging from negative to neutral or even positive across systems (REF). Part of this uncertainty may arise because landscape effects are often inferred from landscape configuration alone, despite connectivity being widely recognized as a fundamental process underlying biodiversity responses to landscape change (REF). Connectivity emerges from the interaction between landscape structure and species dispersal ability (REF). However, empirical studies typically estimate landscape effects by correlating structural landscape metrics with local biodiversity patterns (REF). This leaves the causal link to connectivity implicit, primarily because species dispersal is rarely known in real landscapes (REF).

As a result, there is little consensus on which configuration metrics best represent connectivity or which spatial extents are most relevant for estimating biodiversity responses to landscape change (REF). Consequently, current approaches often infer connectivity-mediated ecological responses from landscape structure without explicitly representing the demographic processes that generate those responses. Resolving this gap is not only a theoretical challenge but also a practical necessity for conservation planning, in which different traditions (from generalizable heuristics to context-specific mechanistic models) offer contrasting yet potentially complementary paths forward (REF).

Specifically, we ask how uncertainty in dispersal limitation influences the magnitude and direction of landscape effects on local tree diversity along an empirical forest-cover gradient. By representing connectivity as an emergent property of dispersal and landscape configuration within a demographic model, we provide a mechanistic basis for disentangling the effects of habitat loss and fragmentation on tropical forest biodiversity.

# Materials and Methods 

## Overview of the mechanistic framework

To investigate how landscape configuration and dispersal limitation jointly influence local tree diversity, we integrated empirical forest inventories, experimental landscapes, and simulations from a Spatially Explicit Neutral Model (SENM) into a single mechanistic framework.
 
Within the SENM, individuals are ecologically equivalent and community dynamics emerge from stochastic processes of birth, death, dispersal, and external species input parameterized by U (REF). Dispersal limitation is represented explicitly by a distance-dependent dispersal kernel parameterized by k, which controls the proportion of propagules retained near parental individuals. Under this formulation, ecological connectivity is treated as an emergent consequence of the interaction between dispersal ability and landscape configuration (REF).

For each empirical forest plot, we constructed three experimental landscapes: (i) the contemporary fragmented landscape, (ii) a clumped landscape in which remaining forest cover was spatially aggregated, and (iii) a pristine continuous-forest landscape representing the absence of habitat loss. Comparing simulation outcomes across these counterfactual landscapes allowed us to isolate the effects of forest cover loss, fragmentation per se, and their combined effects on local diversity.

We repeated simulations across multiple values of k, ranging from strongly to weakly dispersal-limited regimes, to account for uncertainty in dispersal ability. For each regime, the relevant spatial scale was determined by evaluating the sensitivity of U to increasing landscape extent in the pristine scenario. The minimum stabilizing extent then defined the simulation extent for that regime.

Using these extent-specific configurations, the SENM was calibrated independently for each empirical plot and experimental landscape by adjusting U to match simulated local richness to observed richness. Variation in the estimates across experimental landscapes was interpreted as a measure of vulnerability to landscape modification, since U represents the dependence of local diversity on external biodiversity sources.

Finally, the empirical plausibility of the SENM was evaluated by testing whether simulated dynamics reproduced empirical species abundance distributions (SADs). Landscape effects were assessed only for simulations satisfying SAD congruence criteria. This validation step provides empirical support for interpreting simulated experimental contrasts as ecologically plausible landscape effects.

	 	
	 	
## Study system and empirical data
We used the Atlantic Forest as a model system for investigating fragmentation effects in tropical forests. This biome retains less than 30% of its original forest cover, distributed as a mosaic of fragments that provides a broad empirical gradient of forest cover, from near-continuous to highly deforested landscapes, while naturally coupling habitat loss with configuration change (REF). To isolate their independent effects, we therefore required an experimental approach capable of separating cover loss from fragmentation per se.
Tree community data were obtained from the TreeCo database, which compiles phytosociological studies across the Atlantic Forest (@de_lima_how_2015). We selected forest inventories that: (i) sampled trees with DBH ≥ 4.8 cm; (ii) used single plots ≥ 1 ha; and (iii) had high taxonomic reliability and precise geographic coordinates. Landscape structure was characterized using MapBiomas Collection 6 land cover maps (@mapbiomas), with forest cover defined as native forest formations including secondary successional stages. For each plot, we extracted a 16 × 16 km landscape window centered on the plot location, an extent chosen to exceed typical fragmentation-study buffers and to ensure stabilization of model dynamics across all dispersal regimes.

## Experimental design: counterfactual landscapes
To disentangle the effects of forest cover loss and fragmentation per se, we constructed three experimental landscapes for each empirical plot (Fig. S1): (i) the **contemporary fragmented landscape** representing observed forest cover and configuration; (ii) a **pristine continuous-forest landscape** representing no habitat loss; and (iii) a **clumped landscape** in which all remaining forest cover from the fragmented landscape was spatially aggregated into a single square block centered on the plot, preserving total habitat amount while minimizing fragmentation. Comparing simulation outcomes across these landscapes allowed us to isolate three distinct effects: total effect of forest loss and fragmentation (pristine vs. fragmented), fragmentation per se (clumped vs. fragmented), and forest cover loss independent of configuration (pristine vs. clumped). This design resolves a key inferential problem in fragmentation research: because habitat removal inevitably alters configuration in empirical landscapes, complete orthogonality between amount and fragmentation is only achievable through controlled counterfactual manipulation.

## Spatially explicit neutral model
We used a Spatially Explicit Neutral Model (SENM; @rosindell2008 , REF) as a mechanistic framework linking landscape structure to local diversity. The SENM assumes ecological equivalence among individuals, with community dynamics emerging from stochastic birth, death, dispersal, and external species input. Two parameters are central to our analysis. First, **U** represents the probability that an individual lineage originates from a newly introduced species (i.e., external input). Second, **k** quantifies dispersal limitation: the expected proportion of propagules retained near parental individuals. Under this formulation, connectivity emerges from the interaction between dispersal ability (k) and landscape configuration. This property makes the SENM particularly suited for investigating how dispersal limitation modulates biodiversity responses to landscape change (Campos et al. 2012, 2013; Thompson et al. 2019, REF).


## Calibration and validation
We calibrated the SENM independently for each empirical plot, dispersal regime, and experimental landscape by adjusting U so that simulated species richness matched observed richness (semi-analytical method; \cite{rosindell2008}). To account for uncertainty in dispersal ability, we repeated simulations across 20 values of k ranging from strongly (k = 0.99) to weakly (k = 0.05) dispersal-limited regimes. For each regime, we first determined the minimum spatial extent required for U to stabilize in pristine landscapes. This extent was then used for all simulations under that dispersal regime.
Before interpreting landscape effects, we validated whether simulated communities reproduced empirical species abundance distributions (SADs) using Kolmogorov-Smirnov tests with bootstrap resampling. We retained only simulation batches in which at least 75% of replicate SADs were congruent across all three experimental landscapes simultaneously. Furthermore, we restricted interpretation to dispersal regimes showing consistently high mean congruence, thereby empirically delimiting the range of dispersal conditions under which the SENM provided an adequate approximation of the empirical system.
## Vulnerability metric
Within the SENM framework, U measures the dependence of local diversity on external biodiversity sources required to maintain richness under ecological drift (REF). For a given plot, larger U indicates greater reliance on external input (e.g., long-distance dispersal, propagule banks, speciation). Therefore, differences in U between experimental landscapes provide a mechanistic measure of biodiversity vulnerability associated with landscape modification.

For each plot and dispersal regime, we calculated vulnerability contrasts as log-ratios of U estimates:
- Total effect (FLF) = log(U_pristine / U_fragmented)
- Fragmentation _per se_ (FPS) = log(U_clumped / U_fragmented)
- Forest cover loss (LFC) = log(U_pristine / U_clumped)

These three contrasts decompose the total landscape effect additively: FLF = FPS + LFC. The clumped landscape thus serves as an intermediate reference, sharing habitat amount with the fragmented landscape and configuration with the pristine landscape.

## Statistical analyses
We analyzed variation in SENM congruence (proportion of congruent SADs) using Hierarchical Generalized Additive Models (HGAMs) with binomial errors. The most complex model included smooth terms for dispersal limitation (k) with landscape-specific smoothers, a plot-level random effect represented by a spline of k nested within plots, and a bivariate smoother for plot geographic coordinates to account for spatial autocorrelation.
For each vulnerability metric (FLF, FPS, LFC), we fitted a series of Gaussian HGAMs evaluating the influence of remaining forest cover (%FC) and dispersal limitation (k). Models ranged from simple plot intercepts to tensor-product interactions between k and %FC. Model fitting used restricted maximum likelihood (REML; R package mgcv; \cite{wood2017}). Predictions from the best-supported models were used to estimate mean vulnerability responses across observed gradients of dispersal limitation and forest cover.
Detailed procedures are provided in the Supporting Information.

# Results

<p> <!--Resultados auxiliares: exposição e citação do SI -->
Of the forest inventories available in the TreeCo database, 109 met the initial selection criteria (Table SI1). After excluding two highly disturbed plots and two sites for which landscape centering was not feasible, the final dataset comprised 105 plots (Fig. S1). These plots span a broad geographic range across the Atlantic Forest and capture substantial variation in forest structure, including plot area (median = 1 ha, IQR = 0 ha, min = 1 ha, max = 102 ha), abundance (median = 1,528 individuals, IQR = 608, min = 649, max = 12,105), and species richness (median = 107 species, IQR = 66, min = 26, max = 458). All plots were associated with forest cover maps derived from MapBiomas Collection 6. This dataset provides the empirical basis for the neutral simulations. Detailed landscape maps for each plot are presented in Figures SI1.X and SI1.X.
</p>

<p> <!--Calibração e sensibilidade: escala suficiente -->
In pristine landscape simulations at the largest spatial scale, U estimates varied substantially across dispersal scenarios and plots (Fig. S2.X). Estimates of the sufficient spatial scale were more consistent when using a 75% threshold of the total reduction in U (Fig. S2.XX). Under moderate dispersal limitation, uncertainty in the estimation of the sufficient spatial scale increased (Fig. S2.XXX). This occurred because variation across spatial scales was smaller than the variability among simulation replicates for a given parameter combination. Using the 75% threshold and linear interpolation, a 2-km landscape was sufficient for scenarios in which the fraction of propagules remaining within the parental neighbourhood was k ≥ 0.35 (Fig. S2.Y). Under weaker dispersal limitation, the sufficient spatial scale increased to 4 km.
</p>


## Dispersal limitation constrains the empirical validity of a spatially explicit neutral model {Congruencia}


<p> <!--Descrição do padrão empírico chave para o critério post-hoc -->
Mean congruence exceeded 75% in most simulations but declined under very weak dispersal limitation, especially in clumped and pristine landscapes (Fig. 1). There is substantial variability between plot-level trends, with non-linear congruence responses as a function of dispersal-by-landscape interaction (Fig. 1). Figure 1 shows the prediction from the most plausible statistical model, that describes the probability of congruence by the interaction of dispersal and landscape configuration at the plot-level in addition with the spatial proximity between forest plots (@ref(CongSAD)). 
</p>


<p> <!--a importância da interação entre k e landscape-->
Although the most plausible model included both the dispersal-by-landscape interaction per plot and spatial proximity, a simpler model excluding spatial coordinates explained an identical proportion of deviance (94%), had a moderately lower AICc (ΔAICc = 8.79; Table S3.2), and showed no spatial autocorrelation in residuals (Fig. S3.1). In contrast, models that omitted the plot-level interaction exhibited significant positive spatial autocorrelation and a substantial drop in explained deviance (64% and 59%; Table S3.2). This indicates that the plot-level interaction between dispersal and landscape configuration is sufficient to capture the main congruence patterns, with spatial proximity playing a refinining role.
</p>

<p> <!--Critério post-hoc-->
Congruence declined under high dispersal scenarios (Fig. 1), indicating that MNEE's mechanistic assumptions are most plausible when dispersal limitation is more severe. We therefore restricted subsequent analyses to scenarios with moderate-to-strong dispersal limitation (k ≥ 0.50), where MNEE consistently showed high mean congruence across all experimental landscapes. Combining this post-hoc criterion with the predefined 75% congruence threshold across all three landscapes yielded 66 selected plots (62,8%) for the vulnerability metric analysis. Selected and non-selected plots exhibited similar environmental characteristics and were distributed across the Atlantic Forest with no evidence of systematic spatial patterning (Fig. S3.2; Appendix S3).
</p>


## Forest cover and dispersal jointly shape vulnerability to forest loss and fragmentation

Differences in U between landscapes reflect variation in the external replenishment required to sustain observed diversity patterns. Accordingly, log(Ui/Uj) quantifies landscape‑induced vulnerability across three experimental contrasts: total effect of forest loss and fragmentation (FLF; pristine vs. fragmented), fragmentation per se (FPS; clumped vs. fragmented), and forest cover loss (LFC; pristine vs. clumped).

Model selection indicated that the interaction between remaining forest cover (%FC) and dispersal limitation (k) improved model fit for FLF and LFC, but not for FPS (Table 1). For FPS, the best-supported model included only additive effects of %FC and k (plus plot-level random dispersal effects). Notably, across all three landscape effects, accounting for plot-specific dispersal responses was essential to achieve high explanatory power (>75% deviance explained); removing this component reduced deviance explained by 25 to 40 percentage points (Table 1).

The LFC effect became increasingly negative in plots with reduced %FC and was negligible in plots with high %FC (Fig. 2). Across dispersal regimes, LFC exhibited a non-linear response to dispersal limitation (k), with minimum vulnerability at intermediate values (k ≈ 0.70) and stronger effects under both strong and weak dispersal limitation. This pattern indicates that habitat loss exerts its strongest influence when dispersal limitation is either strong or weak, whereas intermediate dispersal partially buffers vulnerability. 

The FPS effect was generally weak, with most values clustering around the null reference in landscapes with forest cover above 60%, and becoming weakly negative at low forest cover (<30%) (Fig. 2). Only a few plots showed plot-specific patterns, with some displaying pronounced negative effects under moderate-to-mild dispersal limitation and others showing weak positive effects under severe limitation. Although the magnitude of FPS occasionally increased in plots with reduced %FC, the most pronounced effects were observed in a small number of plots with intermediate %FC. Thus, fragmentation per se acts primarily as a context-dependent modifier of vulnerability rather than a dominant driver.

The FLF effect became more negative as %FC decreased, although a few plots displayed patterns resembling FPS at intermediate-to-high %FC (Fig. 2). Overall, FLF exhibited greater variability, integrating features of both LFC and FPS. 

A complementary analysis showed that differences in magnitude between LFC and FPS were generally small and often fell within the null reference range, although LFC tended to dominate in plots with reduced %FC (Fig. S4.1). Across plots, LFC and FPS frequently differed in sign (41-65%), more often than they shared the same sign (38–58%). The proportion of cases in which LFC and FPS differed in sign increased with %FC. When both effects were negative, FLF was correspondingly more negative. In contrast, when the effects differed in sign, positive FPS partially counterbalanced negative LFC. 

These contrasting regimes were most evident in low-cover landscapes, where LFC dominated, and in a small number of intermediate-cover plots, where FPS effects prevailed (Fig. 2; Fig. S4.1). Together, these patterns indicate that the combined effect of forest loss and fragmentation is predominantly negative in low-cover landscapes, while in higher-cover landscapes it may reflect plot-specific effects of spatial configuration.


# Discussion

Connectivity is widely recognized as a key mechanism linking landscape structure to biodiversity responses, yet its demographic basis remains difficult to quantify (\cite{swift2010critical}; \cite{villard_beyond_2014}; \cite{rivaEbanskleite2024} REF). Using a spatially explicit neutral model calibrated and validated against empirical tree communities (\cite{rosindell2012case}; \cite{Marshall2014}; \cite{DAGinfEPI2019}; \cite{gianluca2022CH5}), we found that model congruence depended strongly on dispersal limitation and landscape configuration. Within dispersal regimes with high mean congruence, the effects of forest loss and fragmentation showed structured responses across gradients of dispersal limitation and forest cover.

## High SENM mean congruence with empirical SADs fails only under mild dispersal limitation

Except for mild dispersal limitation, the SENM achieved high mean congruence with empirical SADs across a broad range of dispersal and landscape combinations, although congruence varied substantially among plots. This result suggests that relatively simple demographic models can reproduce important features of fragmented tree communities when dispersal limitation and landscape structure are represented explicitly (REF, the case of neutral theory, etc). Such findings contrast with earlier skepticism regarding the ability of neutral models, including spatially explicit formulations, to reproduce biodiversity patterns in fragmented tropical forests (Gilbert et al. 2006; Laurance 2008; Rosindell & Etienne 2011). A key difference concerns the interpretation of the species-input parameter U. Previous simulation studies often treated U primarily as true speciation (REF), which implies low values of U and consequently slow community dynamics because species persistence times increase as U decreases (REF). In our framework, U represents the aggregate contribution of processes introducing novel lineages into local communities, including long-distance dispersal, propagule-bank replenishment, and speciation (REF). This broader interpretation yields higher values of U and faster turnover, potentially explaining the congruence observed across most simulations. Importantly, this hypothesis derives from matching observed SADs rather than temporal community dynamics (REF). 


## Forest cover loss consistently increases vulnerability in low-cover landscapes

Within our framework, U represents the external input required to maintain observed local diversity under ecological drift. Consequently, differences in U between experimental landscapes provide a mechanistic measure of how landscape modification alters the dependence of local patterns on external biodiversity sources. Under this interpretation, forest cover loss was consistently associated with increased vulnerability in low-cover landscapes, whereas its effects were negligible in landscapes with high remaining forest cover (REF galera dos efeitos). The magnitude of this effect was lowest under intermediate dispersal limitation, mirroring the non-linear relationship between dispersal and U observed in pristine landscapes. Under strong dispersal limitation, recruitment is dominated by nearby conspecifics, reducing stochastic species loss and therefore the need for external replenishment (May et al. 2012). As dispersal increases, heterospecific replacement becomes more frequent, increasing drift-driven species loss and raising U. Under weak dispersal limitation, however, distant populations increasingly contribute propagules to local communities, reducing local extinction and causing U to decline. This mechanism may also explain why vulnerability to habitat loss was minimized under intermediate dispersal limitation. Because U is highest under these conditions, local diversity becomes less dependent on surrounding habitat as a source of biodiversity replenishment, weakening the ecological consequences of habitat loss (Campos et al. 2012, 2013).

## Fragmentation per se is generally weaker and highly contingent

In contrast to forest cover loss, fragmentation per se exhibited weak and highly variable effects across landscapes (REF galera da fahrig). The best-supported statistical model included only additive effects of dispersal limitation and forest cover, indicating that fragmentation responses were not organized by a consistent dispersal-by-forest-cover interaction, unlike the pattern observed for habitat loss. Positive and negative FPS effects occurred across both dispersal and forest-cover gradients, with some plots showing positive responses under strong dispersal limitation and others exhibiting strongly negative responses at intermediate to high forest cover. This pattern is broadly consistent with the long-standing inconsistency of fragmentation responses reported across empirical systems (REFs). Fragmentation per se has been associated with positive, neutral, and negative responses depending on the study (REFs). Our results suggest that this variability emerges because fragmentation effects depend on how particular landscape configurations modify effective connectivity under different dispersal constraints. Under severe dispersal limitation, some spatial configurations may favor local conspecific replacement, reducing stochastic species loss and potentially generating positive fragmentation effects in a subset of landscapes (Campos et al. 2012, 2013). Conversely, other configurations may further restrict propagule exchange among local communities, amplifying biodiversity loss (REF). The scale-sensitivity analysis further suggests that these connectivity dynamics depend not only on dispersal limitation and landscape configuration but also on local community properties. Species richness and individual density influence the baseline value of U required to maintain local diversity and therefore affect how strongly changes in landscape configuration alter community dynamics. Consequently, fragmentation effects may emerge from the interplay among spatial configuration, dispersal processes, and local demographic conditions, helping explain why responses remain difficult to generalize across landscapes.

## Forest loss dominates, but fragmentation shapes variability in the total forest loss and fragmentation effect

The total effect of forest loss and fragmentation (FLF) partially reflected the patterns observed for forest cover loss in highly deforested landscapes. Consistent with this result, FLF was best explained by the interaction between dispersal limitation and forest cover, indicating that habitat loss remained the primary determinant of vulnerability across landscapes (REF). However, FLF also inherited the variability associated with fragmentation _per se_. In most low-cover landscapes, negative FPS effects amplified the vulnerability associated with habitat loss, producing stronger total effects (REF). In a smaller number of cases, positive FPS effects partially compensated for habitat-loss vulnerability, particularly under severe dispersal limitation. At intermediate and high forest cover, where LFC effects were weak, variation in FLF reflected the contingent responses associated with fragmentation per se. Together, these results suggest that habitat loss primarily determines the direction of the total landscape effect, whereas fragmentation per se contributes to its variability (REF).

## Implications for the fragmentation debate

Our results revealed a general tendency for forest cover loss to dominate landscape effects in highly deforested landscapes, with fragmentation per se contributing with additional variability. However, the most notable pattern was the substantial variability among plots. Responses to dispersal limitation differed markedly, particularly for fragmentation per se and the total effect of forest loss and fragmentation. Consistent with this result, statistical models that ignored plot-level variation in dispersal responses showed substantially lower explanatory power. Although the SENM did not provide an equally good description of all communities, the framework nevertheless allowed landscape effects to be estimated at the plot level (REF). This contrasts with many observational approaches, where inference is typically restricted to average effects across plots (REF). From this perspective, explaining variation among communities may be as important as estimating average landscape effects.

This distinction reflects a broader tension within landscape ecology regarding how biodiversity responses should be incorporated into conservation planning. One tradition emphasizes generalizable heuristics derived from broad comparative analyses of landscape structure and biodiversity responses, prioritizing scalable recommendations based on relatively simple landscape descriptors (e.g., Fahrig 2017; Watling et al. 2020; Arroyo-Rodríguez et al. 2020). In contrast, more mechanistic approaches seek to represent species responses, dispersal dynamics, and restoration scenarios explicitly through spatially detailed models tailored to particular ecological contexts (Banks-Leite et al. 2021). More broadly, these traditions reflect different strategies for balancing ecological realism, inferential generality, and data availability in conservation science.

The SENM framework provides a possible integration between these approaches by explicitly incorporating dispersal limitation and landscape configuration while remaining parameterizable using biodiversity and landscape data that are broadly available across fragmented systems. In contrast to spatially implicit approaches based primarily on habitat amount or species–area relationships (e.g., Strassburg et al. 2019), the framework allows alternative connectivity scenarios to be evaluated explicitly through counterfactual landscape simulations. This framework may help reconcile the need for mechanistic ecological interpretation with the practical demand for scalable inference in data-limited conservation contexts.

## Limitations

### Orthogonality and decomposition of landscape effects

Within our framework, the clumped scenario allows the effects of spatial aggregation to be evaluated independently from overall forest cover (REF Fahrig et al. ). This formulation also clarifies an inferential problem underlying the fragmentation debate. Once habitat is spatially distributed across a landscape, habitat removal inevitably alters the spatial configuration of the remaining habitat (REF). In this sense, the fragmentation debate may be interpreted less as a disagreement about whether fragmentation matters and more as a question about the conditions under which total landscape effects can be meaningfully decomposed into independent ecological components (Fahrig 2003, 2019; Didham et al. 2012; Fletcher Jr et al. 2018; Fahrig et al. 2019; Püttker et al. 2020). Complete orthogonality between habitat amount and fragmentation therefore represents a limiting inferential case rather than an intrinsic property of empirical landscapes (Pearl et al. REF).

Critically, constructing the clumped landscape depends on defining the spatial extent from which habitat is redistributed around the focal plot, making scale specification an intrinsic component of the decomposition between forest cover loss and fragmentation per se (REF). This requirement differs from the estimation of the total effect of forest loss and fragmentation, which only requires a landscape extent large enough to avoid constraining dispersal processes, once the SENM can perform simulations in infinite landscapes (REF). Consequently, decomposing landscape effects imposes stronger inferential requirements than estimating their combined effect, particularly in data-limited conservation contexts where independent information for defining ecologically relevant spatial scales is often unavailable (REF). Under these conditions, inference depends increasingly on auxiliary assumptions linking contemporary biodiversity patterns to landscape structure (REF).


### Temporal correspondence between biodiversity and landscape patterns

Most observational studies relate contemporary communities to contemporary landscape configuration, implicitly assuming that the landscape characteristics relevant to community assembly have remained sufficiently stable through time or that delayed ecological responses have largely relaxed (Frazier & Kedron 2017; REF). Violations of this assumption may contribute substantially to inferential uncertainty, but historical information on habitat dynamics is often unavailable to assess their magnitude (REF).

Within the SENM framework, this assumption is operationalized through ecological equilibrium (REF). Under this interpretation, biodiversity patterns are assumed to reflect the long-term interaction between dispersal processes and landscape structure rather than transient responses to recent environmental change (REF). Consequently, deviations between observed and simulated communities may partly reflect temporal mismatches between contemporary biodiversity patterns and the landscape conditions used to model them.

Nevertheless, our results suggest that contemporary landscape structure may retain information about the processes shaping present-day communities. The interaction between dispersal limitation and landscape configuration was sufficient to account for variation in model congruence among plots and eliminate positive spatial autocorrelation in residuals. Although this result does not directly evaluate historical landscape dynamics, it suggests that contemporary landscape structure contains information relevant to explaining both biodiversity patterns and their spatial organization, even in the absence of explicit historical data (REF).


## Future directions

Incorporating dynamic landscape change represents another important avenue for future research (REF). Explicitly modeling habitat regeneration and landscape transitions could improve the representation of temporal connectivity and delayed ecological responses (REF). However, such extensions would substantially increase both data requirements and computational complexity. The current coalescent implementation reconstructs equilibrium communities directly, allowing large-scale analyses from relatively limited biodiversity and landscape data (REF). Dynamic landscape models would require additional information on landscape trajectories and transient community dynamics, potentially reducing the scalability that constitutes one of the main practical advantages of the present framework.

In contrast, some additional ecological mechanisms may extend the present framework without fundamentally altering its inferential structure. Matrix effects, for example, may influence biodiversity dynamics by modifying dispersal success among habitat patches and demographic processes near habitat edges (REF). Representing these mechanisms explicitly within the same counterfactual framework may help distinguish the relative contributions of forest loss and fragmentation from those associated with matrix. Unlike the decomposition of habitat loss and fragmentation per se, matrix effects can be implemented directly as local dispersal and demographic processes (REF), reducing dependence on a priori definitions of spatial extent.


```{r caption tab-selecao,echo=FALSE,include=TRUE,eval=TRUE}
library(officer)
f_z <- function(x) (x-mean(x))/sd(x)
f_label <- \(vn){
  gsub("1","M4",vn) %>% 
    gsub("s\\(k,by\\=land\\) \\+ \\(lat,long\\)","M1",.) %>% 
    gsub("s\\(k,by\\=land\\) - te\\(k,plot,by\\=land\\)","M3",.) %>% 
    gsub("s\\(k,by\\=land\\)","M2",.)
    
}
f_antilabel <- \(vn){
  gsub("M4","1",vn) %>% 
    gsub("M1","s\\(k,by\\=land\\) \\+ \\(lat,long\\)",.) %>% 
    gsub("M3","s\\(k,by\\=land\\) - te\\(k,plot,by\\=land\\)",.) %>% 
    gsub("M2","s\\(k,by\\=land\\)",.)
    
}
vcap <- "Model comparison among candidate hierarchical generalized additive models (HGAMs). ΔAICc represents the difference in AICc relative to the best-supported model; est. coef. denotes the effective degrees of freedom (model complexity); weight indicates Akaike weight; Dev. Exp. represents deviance explained. Model formulas are expressed using the mgcv package syntax: s() denotes a univariate smooth term, and te() denotes a tensor-product smooth for interactions between continuous variables (here, dispersal limitation, k, and landscape forest cover, p). The notation | Plot indicates plot-specific smooth terms (hierarchical structure), allowing dispersal effects to vary among plots."
```
```{r tab-selecaoCong,echo=FALSE,include=TRUE,eval=TRUE}
df_tabsel <- readRDS(file="1_to_compile_dissertacao_EM_USO/09_SI/RDS/tabsel_cong_logUU.rds") %>% 
  mutate(dev.expl=100*dev.expl,
         across(where(is.numeric), ~round(.x,2)),
         Effect=base,
         modelo=gsub("Site","Plot",modelo)) %>% 
  select(-base)
library(flextable)
f_label <- \(vn){
  gsub("1","M4",vn) %>% 
    gsub("s\\(k,by\\=land\\) \\+ \\(lat,long\\)","M1",.) %>% 
    gsub("s\\(k,by\\=land\\) - te\\(k,plot,by\\=land\\)","M3",.) %>% 
    gsub("s\\(k,by\\=land\\)","M2",.)
    
}
f_flextable <- \(dfi){
  #
  df_grouped <- as_grouped_data(dfi, groups = "Effect")
  #
  as_flextable(df_grouped) %>%
    # Estética geral
    bg(bg = "white", part = "all") %>%
    theme_vanilla() %>% 
    # 
    set_header_labels(
      modelo = "HGAM",
      dAICc = "ΔAICc",
      df = "est. coef.", # 
      weight = "weight",
      dev.expl = "Dev. Exp."
    ) %>%
    
    # Destacar as linhas que são os nomes dos grupos (Frag. total, etc)
    bold(i = ~ !is.na(Effect), bold = TRUE, part = "body") %>%
    #as_italic(i = ~ !is.na(base), part = "body") %>%
    bg(i = ~ !is.na(Effect), bg = "#F7F7F7", part = "body") %>% # 
    
    # Formatação de números
    colformat_double(digits = 2) %>% 
    colformat_double(j = "weight", digits = 2) %>% # 
    
    # Alinhamento e legenda
    align(align = "center", part = "all") %>% 
    align(j = 1, align = "left", part = "body") %>% # 
    # set_caption(caption = vcap) %>% 
    add_header_lines(values = as_paragraph(
            as_chunk("Table 1. ", props = fp_text(bold = TRUE)),
            as_chunk(vcap, 
                     props = fp_text(bold = FALSE),
                     fp_p = fp_par(text.align = "justify") 
                     )
             )
            ) %>%
    # 
    set_table_properties(layout = "fixed") %>%
    # Define a largura apenas para a coluna 'modelo'
    flextable::width(j = "modelo", 
                     width = 10.5/5, unit = "in") %>%
    # Define a largura para todas as OUTRAS colunas
    flextable::width(j = c("dAICc", "df", "weight", "dev.expl"), 
                     width = 1, unit = "in")
}
fxtow <- f_flextable(df_tabsel)
saveRDS(fxtow,file = "1_to_compile_dissertacao_EM_USO/09_SI/RDS/fxtow_tab1.rds")
save_as_image(fxtow, 
              path = "figuras/tabsel_efeitos.png")
```
```{r display da tabela 1}
library(knitr)
library(kableExtra)

df_tabsel <- readRDS(file="1_to_compile_dissertacao_EM_USO/09_SI/RDS/tabsel_cong_logUU.rds") %>% 
  mutate(dev.expl=100*dev.expl,
         across(where(is.numeric), ~round(.x,2)),
         Effect=base,
         modelo=gsub("Site","Plot",modelo)) %>% 
  select(-base)

# Aplicar o label para tornar os nomes dos modelos mais legíveis
df_tabsel$modelo <- f_label(df_tabsel$modelo)

# Criar a tabela com kableExtra
kable(df_tabsel, 
      caption = vcap,  # <-- a legenda vai aqui
      booktabs = TRUE,
      linesep = "",
      format = "latex") %>%
  kable_styling(latex_options = c("striped", "hold_position")) %>%
  add_header_above(c(" " = 1, " " = 1, "Model fit criteria" = 4)) %>%
  pack_rows(index = table(df_tabsel$Effect)) %>%  # agrupa por Effect
  column_spec(1, width = "5cm") %>%
  column_spec(2:5, width = "1cm")
```




```{r criação da figCong, eval=FALSE,include=FALSE,echo=FALSE}
l_dfpred <- readRDS(file = "dados/csv_SoE/rds/l_dfpred_md_cong_absoluta.rds")
library(hexbin)
p <- l_dfpred$fixo_e_aleat %>% 
  ggplot(aes(x=k,y=nSAD)) +
  # geom_point(alpha=0.3) +
  geom_line(aes(y=fit,group = SiteCode),alpha=0.3) +
  # geom_boxplot(aes(group=k),alpha=0.6) +
  geom_ribbon(data=l_dfpred$fixo,
              aes(x=k,y=fit,ymin=lower,ymax=upper),
              color="blue",
              fill="lightblue") +
  geom_line(data=l_dfpred$fixo,
            aes(x=k,y=fit),
            color="darkred") +
  geom_hex(bins = 50,alpha=0.5) + 
  scale_fill_gradient("count",low = "gray", high = "black", na.value = NA) +
  scale_x_continuous(expand = c(0.01, 0)) +
  scale_y_continuous(expand = c(0.01, 0)) +
  labs(x="Fraction of propagules within the immediate neighborhood (k)",
       y="Number of congruent simulated SADs (goodness-of-fit)" #,
       # title="Congruence of simulated SADs across experimental landscapes",
       # subtitle="Congruence is best described by connectivity and spatial proximity among forest plots"
       ) +
  facet_wrap(~land) +
  theme_bw() +
  theme(strip.text = element_text(size=12, #margin=margin(),
                                  face="bold"))
saveRDS(p,file = "1_to_compile_dissertacao_EM_USO/09_SI/RDS/p_interpretacao_cong.rds")
ggsave(filename="figuras/descricao_congruencia_absoluta.png",
       p,height = 5,width = 13.8)
library(magick)
img <- image_read("figuras/descricao_congruencia_absoluta.png") %>% 
  image_resize("50%") %>% 
  image_trim()
image_write(img,path = "figuras/descricao_congruencia_absoluta.png")
```
```{r captio figCong,eval=TRUE,echo=FALSE}
vcap <- "Congruence between simulated and observed SADs (y-axis) across dispersal (x-axis) and landscape combinations (facet labels). Hexagonal bins show the density of plot‑level observations (lighter = fewer points, darker = up to 100 points). Gray lines: plot‑level predictions from the most plausible model with spatial coordinates set to zero. Red line: overall mean prediction; blue shading: 95% pointwise confidence interval. Mean congruence exceeded 75% in most simulations but declined under very weak dispersal limitation, especially in clumped and pristine landscapes."
```
```{r figCong,fig.cap=vcap,fig.width=12,fig.height=7,eval=TRUE,include=TRUE}
p <- readRDS(file = "1_to_compile_dissertacao_EM_USO/09_SI/RDS/p_interpretacao_cong.rds")
p
```

```{r criacao de figura pefeitos,eval=FALSE,include=FALSE,echo=FALSE}
library(metR)
library(gridExtra)
library(ggplot2)
library(dplyr)
#-------------------------------------1----------------------
#--------------------- base de dados  ----------------------
#-----------------------------------------------------------
df_real <- readRDS("dados/csv_SoE/rds/df_obs_pred_plot_logUU_pk.rds") %>% 
  mutate(
    p_class = case_when(
       p>=0.99 ~ "%FC ≥ 99",
       p<0.99 & p>=0.80 ~ "80 ≤ %FC < 99",
       p<0.80 & p>=0.60 ~ "60 ≤ %FC < 80",
       p<0.60 & p>=0.30 ~ "30 ≤ %FC < 60",
       p<0.30 ~ "%FC < 30"),
    p_class = factor(p_class,levels=c("%FC < 30",
                                           "30 ≤ %FC < 60",
                                           "60 ≤ %FC < 80",
                                           "80 ≤ %FC < 99",
                                           "%FC ≥ 99")),
    lower = fit - 1.96 * se.fit,
    upper = fit + 1.96 * se.fit
  )
#-----------------------------------------------------------
#------------------ função da figura -----------------------
#-----------------------------------------------------------
f_obs_predito_bysite <- \(dff,logic_ribbon=FALSE,vpos=c(0.72,0.25),vt=15,vts=0.90){
  dfpred <- dff %>% 
    mutate(
      p_class = case_when(
        p>=0.99 ~ "%FC ≥ 99",
        p<0.99 & p>=0.80 ~ "80 ≤ %FC < 99",
        p<0.80 & p>=0.60 ~ "60 ≤ %FC < 80",
        p<0.60 & p>=0.30 ~ "30 ≤ %FC < 60",
        p<0.30 ~ "%FC < 30"),
      p_class = factor(p_class,levels=c("%FC < 30",
                                        "30 ≤ %FC < 60",
                                        "60 ≤ %FC < 80",
                                        "80 ≤ %FC < 99",
                                        "%FC ≥ 99")),
      lower = fit - 1.96 * se.fit,
      upper = fit + 1.96 * se.fit
    )
  df_ij <- dfpred %>% 
    select(SiteCode,p_class) %>% 
    distinct() %>% 
    group_by(p_class) %>% 
    tally() %>% 
    mutate(label = paste0(p_class,", n = ",n)) %>% 
    select(-n)
  df_ij$label <- factor(df_ij$label,
                        levels=unique(df_ij$label))
  dfp <- inner_join(dfpred,df_ij)
  nsites <- filter(dfp,p_class=="%FC ≥ 99") %>% pull(SiteCode) %>% unique %>% length()
  dfp <- mutate(dfp,
                efeito = gsub("FLF","Forest Loss and Frag. (FLF)",efeito) %>% 
                  gsub("FPS","Frag. per se (FPS)",.) %>% 
                  gsub("LFC","Loss of Forest Cover (LFC)",.)
                )
  dfrange <- filter(dfp,p_class=="%FC ≥ 99") %>% 
    reframe(
      names = c("amplitude","amplitude","Q 5%-95%","Q 5%-95%"),
      value = c(min(Uefeito), max(Uefeito),quantile(Uefeito,probs=0.05),quantile(Uefeito,probs=0.95))
    )
  dfp <- filter(dfp,p_class!="%FC ≥ 99")
  p <- dfp %>% 
    mutate(efeito = factor(efeito,
                           levels=c("Forest Loss and Frag. (FLF)",
                                    "Frag. per se (FPS)",
                                    "Loss of Forest Cover (LFC)"))) %>% 
    ggplot(aes(x=k,y=Uefeito)) +
    geom_boxplot(aes(group=k),alpha=0.6) +
    geom_hline(yintercept = 0,color="black") +
    geom_hline(data = dfrange, aes(yintercept = value,linetype=names),
               color="black",linewidth=0.5) +
    scale_linetype_manual(name = paste0("%FC ≥ 99",", n = ",nsites),
                          values=c("dotted","dashed")) +
    {if(logic_ribbon)geom_ribbon(aes(x=k,y=fit,ymin=lower,ymax=upper,group = SiteCode),
                                 fill="lightblue",
                                 alpha=0.2)}+
    geom_line(aes(y=fit,group = SiteCode,color=p),alpha=0.4) + #,color="darkred"
    geom_point(aes(color=p),alpha=0.85) +
    scale_colour_gradient2("% FC",midpoint=0.5,
                           low="red",
                           mid = "yellow",
                           high = "darkgreen") +
    labs(x="Fraction of propagules dispersing locally (k)",y="logU/U",title="") +
    facet_grid(efeito~label,scales="free") +
    theme_linedraw() +
    theme(plot.margin=unit(c(0,0.2,0,0), "cm"),
          text=element_text(size=vt,face="bold"),
          strip.background = element_rect(fill = "white"),
          strip.text = element_text(color="black"),
          legend.position = "inside",
          legend.position.inside = vpos,
          legend.direction="horizontal",
          legend.box = "horizontal",
          legend.box.just = "left",
          legend.background = element_rect(color="black")
          ) +
    guides(linetype = guide_legend(
      title.theme = element_text(size = vt*vts, face = "bold")
      )
    )
  return(list("plot"=p,"range"=dfrange))
}
#
#-----------------------------------------------------------
#------------------ criação da figura ----------------------
#-----------------------------------------------------------
lret <- f_obs_predito_bysite(df_real,logic_ribbon = TRUE, vpos=c(0.72,0.05),vt = 12.5)
p_efeitos <- lret$plot
vfac <- 1.2
ggsave(filename = "1_to_compile_dissertacao_EM_USO/00_Resultados/figuras/p_efeitos_porclasseCF.png",
       dpi=200,
       plot=p_efeitos,
       width=12*vfac,height=9*vfac)
saveRDS(p_efeitos,file="1_to_compile_dissertacao_EM_USO/00_Resultados/figuras/p_efeitos_porclasseCF.rds")
dfrange <- lret$range
saveRDS(dfrange,file="1_to_compile_dissertacao_EM_USO/00_Resultados/figuras/dfrange.rds")
```
```{r}
library(glue)
dfrange <- readRDS(file="1_to_compile_dissertacao_EM_USO/00_Resultados/figuras/dfrange.rds")
vlabel <- sapply(dfrange$value,round,digits=3)
names(vlabel) <- paste0(rep(c("neg","pos"),2),dfrange$names)
vcap <- glue("Proposed landscape effects described by the functional metric log Ui/Uj, shown as observed values and predictions from a hierarchical generalized additive model (HGAM). Differences in U reflect variation in the external replenishment required to sustain observed diversity patterns. The metric log(Ui/Uj) quantifies the relative change in this requirement and can thus be interpreted as a measure of landscape-induced vulnerability. Each point represents the comparison between two simulations, varying according to the landscape effect under evaluation (row titles). Forest Loss and Fragmentation (FLF): pristine vs  fragmented (joint change in forest amount and configuration); Fragmentation per se (FPS): clumped vs fragmented (change in configuration, constant forest amount); Loss of Forest Cover (LFC): pristine vs clumped (change in forest amount, constant aggregation). The x-axis represents dispersal limitation (k), the fraction of propagules dispersed within the immediate neighbourhood of the parent individual. Points are coloured according to the percentage of forest cover (%FC) observed within the 2-km landscape surrounding each focal plot. Columns correspond to classes of observed %FC for each plot (column titles). Plots with greater and equal 99% forest cover were excluded from the graphical display; only the amplitude (maximum = 0.075; minimum = -0.074) and the central 90% interval (95th quantile = 0.04; 5th quantile = -0.038 ) of log Ui/Uj are indicated on the y-axis. This variability reflects the stochasticity inherent to the spatially explicit neutral model. Boxplots summarise values for each level of k. The y-axis scale differs among landscape effects. Only plots classified as congruent are shown.")
```
```{r p-efeitos, fig.cap=vcap,fig.width=14,fig.height=10,cache=TRUE}
p_efeitos <- readRDS("1_to_compile_dissertacao_EM_USO/00_Resultados/figuras/p_efeitos_porclasseCF.rds")
p_efeitos
```




```{r cap fig 3}
vcap <- ""
```
```{r fig3, fig.cap=vcap, include=FALSE,eval=FALSE,echo=FALSE}
f_ggplot <- \(reldist_points = 0.25,
              size_points,
              reltam_prag = 2/3,
              alpha_eixos_epist = 0.25,
              reltam_epist = 1,
              leg_top=-200,
              legtexts=10,
              relpos_legprag = 0.9,
              vstext=10){
  library(ggplot2)
  # pontos de referência do programa de pesquisa
  df_points <- data.frame(
    x = c(0, 0, 0),
    y = c(0, -1.5, 1.5) * reldist_points,
    tipo = factor(c("origem", "validacao", "invalidacao"),
                  levels=c("origem", "validacao", "invalidacao"))
  )
  # Eixos pragmáticos
  eixo_prag <- data.frame(
    x = c(-3, -3),
    y = c(-1, 1),
    xend = c(3, 3),
    yend = c(1, -1)
  ) * reltam_prag
  eixo_prag_lab <- data.frame(
    x = eixo_prag$xend * relpos_legprag,
    y = eixo_prag$yend,
    label = c("informação", "computações")
  )
  eixo_prag_lab[2,c("x","y")] <- eixo_prag_lab[2,c("x","y")]*(-1)
  # Eixos epistemológicos
  eixo_epist <- data.frame(
    x = c(0, 0),
    y = c(1.5, 1.5),
    xend = c(1.5 * cos(pi/6), -1.5 * cos(pi/6)),  # 30° e 150°
    yend = c(1.5 + 1.5 * sin(pi/6), 1.5 + 1.5 * sin(pi/6))
  ) * reltam_epist
  eixo_epist$y <- df_points$y[df_points$tipo=="invalidacao"]
  eixo_epist_lab <- data.frame(
    x = eixo_epist$xend,
    y = eixo_epist$yend,
    label = c("equivalência\nfuncional", "pressupostos\nauxiliares")
  )
  # ggplot2
  ggplot() +
    # padronizações
    geom_segment(aes(x = 0, y = 0, xend = 0, yend = df_points$y[df_points$tipo=="validacao"]), color = "darkgrey", linewidth = 0.5) +
    geom_segment(aes(x = 0, y = 0, xend = 0, yend = df_points$y[df_points$tipo=="invalidacao"]), color = "darkgrey", linewidth = 0.5) +
    coord_fixed(xlim = c(-3.5, 3.5), ylim = c(-2.5, 3)) +
    labs(x = NULL, y = NULL, color = NULL, shape = NULL) +
    theme_void() +
    guides(
      fill = guide_legend(order = 1, ncol = 1),
      ) +
    # espaço pragmático - setas contínuas 
    geom_segment(
      data = eixo_prag,
      aes(x = x, y = y, xend = xend, yend = yend),
      arrow = arrow(length = unit(0.25, "cm")),
      color = "black", size = 1.5
    ) +
    geom_text(
      data = eixo_prag_lab,
      aes(x = x, y = y, label = label),
      hjust = 0.5, vjust = -0.5,
      size = vstext
    ) +
    # espaço epistemológico - setas tracejadas
    geom_segment(
      data = eixo_epist,
      aes(x = x, y = y, xend = xend, yend = yend),
      arrow = arrow(length = unit(0.25, "cm")),
      color = "black", size = 1, linetype = "dashed"
    ) +
    geom_segment(
      data = eixo_epist,
      aes(x = x, y = y, xend = xend, yend = yend),
      arrow = arrow(length = unit(0.25, "cm")),
      color = "black", size = 1, alpha=alpha_eixos_epist
    ) +
    geom_text(
      data = eixo_epist_lab,
      aes(x = x, y = y, label = label),
      hjust = 0.5, vjust = -0.5,
      size = vstext
    ) +
    # referências do programa - pontos
    geom_point(
      data = df_points,
      aes(x = x, y = y, shape = tipo, fill = tipo),
      color = "gray", 
      size = rep(size_points,3)
    ) +
    # Escalas e temas
    scale_shape_manual(
      name = NULL,
      values = c("invalidacao" = 25, "origem" = 21, "validacao" = 24),
      labels = c(
        "invalidacao" = "Invalidado empiricamente: modificar o modelo",
        "origem" = "MNEE",
        "validacao" = "Validado empiricamente: permite estimar efeitos")
      ) +
    scale_fill_manual(
      name = NULL,
      values = c("invalidacao" = "red", "origem" = "black", "validacao" = "green3"),
      labels = c(
        "invalidacao" = "Invalidado empiricamente: modificar o modelo",
        "origem" = "MNEE",
        "validacao" = "Validado empiricamente: permite estimar efeitos")
      ) +
    guides(
      fill = guide_legend(order = 1, ncol = 1, override.aes = list(size = 6)),
      shape = guide_legend(order = 1, ncol = 1, override.aes = list(size = 6))
      ) + 
    # tema
    theme(
      text = element_text(face="bold"),
      legend.position = "bottom",
      legend.text = element_text(size = legtexts,face="bold"),
      legend.box = "vertical",   # garante empilhamento vertical
      legend.title = element_blank(),
      legend.margin = margin(t = leg_top, unit = "pt"),
      plot.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "pt")
    )
}
p <- f_ggplot(
  reldist_points=0.25,
  size_points=15,
  reltam_epist = 0.75,
  reltam_prag = 3/5,
  alpha_eixos_epist = 0.50,
  vstext = 6.5,
  legtexts = 12,
  leg_top=-225,
  relpos_legprag=0.9
  )
vs <- 0.675
rect_grob <- rectGrob(
  x = 0.5-0.005, y = 0.5+0.025, width = vs*(8.25/10), height = vs*(10/11),
  gp = gpar(col = "black", fill = NA, lwd = 5)
)
p_boxed <- p + 
  annotation_custom(rect_grob,
                    xmin = -Inf, xmax = Inf,
                    ymin = -Inf, ymax = Inf)
ggsave(filename="./1_to_compile_dissertacao_EM_USO/05_Discussao/esquema_final.png",
       plot=p_boxed,
       dpi=200,
       width = 12,
       height = 10,
       bg = "white")
img_obj <- image_read("./1_to_compile_dissertacao_EM_USO/05_Discussao/esquema_final.png") %>% 
  image_trim
image_write(image = img_obj,path = "./1_to_compile_dissertacao_EM_USO/05_Discussao/esquema_final.png")
```


```{=latex}
\bibliographystyle{plainnat}
\bibliography{resources/citations2}
```
