---
title: "Complementary Methodology"
author: ""
output:
  bookdown::pdf_document2:
    latex_engine: lualatex
    pandoc_args: [
      "--variable", "geometry=a4paper",
      "--variable", "geometry=twoside",
      "--variable", "geometry=top=3cm",
      "--variable", "geometry=bottom=2cm",
      "--variable", "geometry=inner=3cm",
      "--variable", "geometry=outer=2cm"
      ]
  bookdown::word_document2: default
  pdf_document: default
  html_document:
    df_print: paged
bibliography: resources/citations2.bib
csl: "resources/ecology-letters.csl"
editor_options:
  chunk_output_type: console
header-includes:
  - \usepackage{polyglossia}
  - \setmainlanguage{english}
  - \renewcommand{\contentsname}{Table of Contents}
  - \usepackage{caption}
  - \captionsetup[figure]{name=Figure, font=it}
  - \captionsetup[table]{name=Table, font=it}
  - \usepackage{tcolorbox}
  - \usepackage{hyperref}
  - \hypersetup{colorlinks=true, linkcolor=blue, citecolor=red, urlcolor=blue}
  - \usepackage[brazilian,hyperpageref]{backref}
  - \tcbuselibrary{breakable}
  - \tcbuselibrary{skins}   
  - \usepackage{lmodern}
  - \usepackage{fontspec}
  - \usepackage{threeparttable} 
  - \usepackage{caption}
  - \usepackage{setspace}
  - \usepackage{etoolbox}
  - \setmainfont{Times New Roman}
  - \fontsize{12}{22}
  - \setlength{\parindent}{1.5cm}
  - \onehalfspacing
  - \makeatletter
  - \patchcmd{\@startsection}{\@afterindentfalse}{\@afterindenttrue}{}{}
  - \makeatother
  - \captionsetup[figure]{font=it}
  - \captionsetup[table]{font=it}
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

\appendix
\renewcommand{\thefigure}{SI1.\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{SI1.\arabic{table}}
\setcounter{table}{0}
\renewcommand{\thesection}{SI1.\arabic{section}}
\setcounter{section}{0} 

<!-- # Apêndice 1: Complementary Methodology  {-#SI} -->

```{r setup SI,include=TRUE,eval=TRUE,echo=FALSE,message = FALSE,warning = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      cache=TRUE,
                      fig.pos = "H", 
                      fig.show = "hold",
                      dpi = 72)
# pacotes
# library(dagitty)
# library(ggdag)
library(gt)
library(gratia)
library(sads)
library(doMC)
# library(metR)
library(gridExtra)
library(ggplot2)
library(readr)
library(purrr)
library(stringr)
library(tidyr)
# library(MuMIn)
# library(AICcmodavg)
# library(insight)
library(bbmle)
library(DHARMa)
library(mgcv)
library(lme4)
library(data.table)
library(plyr)
library(dplyr)
library(flextable)
source("source/nameModel.R")
source("source/GAMMtools.R")
source("source/general_tools.R")
f_z <- function(x) (x-mean(x))/sd(x)
#-------- dados
#- dados tif
df_dados <- read_csv("dados/df_dados_disponiveis.csv") |> 
  dplyr::select(SiteCode, effort_ha, Ntotal, S_obs) |> 
  inner_join(data.frame(tif.path = list.files(path="dados/paisagens/paisagens_mapbiomas6_padronizadas",
                                              pattern=".tif",full.names=T),
                        SiteCode = str_extract(list.files(path="dados/paisagens/paisagens_mapbiomas6_padronizadas",
                                                    pattern=".tif",full.names=T),
                                               "(?<=as\\/).*?(?=.tif)")),
             by = "SiteCode")
```

```{r trackdown chunk SI, include=FALSE,eval=FALSE}
# pacotes
library(trackdown)
# both overwrite: 
# overwrite the current google drive file with the current Rmd:
# upload_file(file ="apendices/A2_figuras_tabelas/A2_figuras_tabelas.Rmd",
#             gpath = "mestrado/artigo_principal/apendices/",
#             hide_code = TRUE)
update_file(file ="apendices/A2_figuras_tabelas/A2_figuras_tabelas.Rmd",
            gpath = "mestrado/artigo_principal/apendices/",
            hide_code = TRUE)
# overwrite the current Rmd with the current google drive 
download_file(file = "secoes_texto/Resultados/Resultados.Rmd",
              gpath = "mestrado/artigo_principal/secoes_texto/")
```

# The features from the pre-selected forest plots   {.unnumbered}

Of the forest inventories available in the TreeCo database, 109 met the initial selection criteria (Table \@ref(tab:dados-disponiveis)). The sampled plots were distributed between 31°S and 7°S latitude and 55°W and 35°W longitude (Fig. \@ref(fig:GE-sitios)a). Most inventories were conducted in areas classified as primary forests according to TreeCo (@de_lima_how_2015). Only two plots, identified as having very high disturbance, were removed from the analysis (Fig. \@ref(fig:GE-sitios)c), resulting in 107 valid inventories.

These plots have a median sampled area of 1 ha, with a median of 1,540 individuals and 107 recorded species. Sampling years ranged from 1986 to 2016 (Fig. \@ref(fig:GE-sitios)b). All 107 plots were associated with forest cover maps derived from MapBiomas Collection 6 (@mapbiomas). However, in two cases, it was not possible to center the square plot within the landscape due to the spatial configuration of the forest cover, reducing the final set of sites to 105 plots (Fig. @ref(fig:sitios-removidos)). Detailed landscape maps for each plot are available in Figures SI3.5 and SI3.6.

```{r caption tabela caracteriscas das parcelas}
vcap <- "Characteristics of the forest plots used in the analysis. Columns represent: Plot Code (unique site identifier); Lat and Long (geographic coordinates); Plot Area (total sampled area in hectares); Inventory criteria (minimum inclusion limit for tree diameter, where DBH is Diameter at Breast Height and PBH is Perimeter at Breast Height); N (total number of individuals); S (observed species richness); Forest Type (phytophysiognomy classification); Pert. class (disturbance history); and Data Year (year of sampling or closest proxy). Disturbance classes represent a combination of intensity and time since the last major event: 'low' (no known disturbance in the last 80 years); 'medium' (moderate disturbance between 80 and 50 years ago); 'high' (medium to high disturbance within the last 50 years); and 'very high' (medium to high disturbance within the last 20 years)."
```
```{r dados-disponiveis, tab.cap=vcap,warning=FALSE,message=FALSE,tab.width=12,results='asis'}
library(readr)
library(dplyr)
select <- dplyr::select
rename <- dplyr::rename
df_dados_disponiveis <- read_csv(file = "~/Documentos/mestrado_Ecologia/artigo_principal/dados/df_dados_disponiveis.csv") %>% 
  mutate(LAT = ifelse(is.na(lat_correct),lat,lat_correct),
         LONG = ifelse(is.na(long_correct),long,long_correct),
         across(where(is.numeric),~round(.x,digits=2)),
         pert_class = factor(forest_succession,
                             levels=c("primary",
                                      "primary/secondary",
                                      "secondary",
                                      "capoeira"),
                             labels=c("low",
                                      "medium",
                                      "high",
                                      "very high")),
         forest_type_label = gsub("Formacao pioneira sob influencia marinha",
                                  "FPIM",
                                  forest_type) %>% 
           gsub("Contato ","",.),
         year_bestProxy = gsub("2012_2013","2012\n-2013",year_bestProxy)
         ) %>% 
  select(SiteCode, LAT, LONG, effort_ha, dbh_cutoff, Ntotal, S_obs, forest_type_label,
         pert_class, year_bestProxy)
##############################################
library(flextable)
library(dplyr)

# 1. Preparar os dados (opcional, ou faça direto no set_header_labels)
ft_desc <- flextable(df_dados_disponiveis) |>
  set_header_labels(
    SiteCode = "Plot Code",
    LAT = "Lat",
    LONG = "Long",
    effort_ha = "Plot Area (ha)",
    dbh_cutoff = "Inventory criteria",
    Ntotal = "N",
    S_obs = "S",
    forest_type_label = "Forest Type",
    pert_class = "Pert. class",
    year_bestProxy = "Data Year"
  ) |>
  # 2. Configurações para o PDF não vazar
  set_table_properties(layout = "fixed") |>
  # Ajuste as larguras (em polegadas) para somar ~6.3
  width(width = c(0.7, 0.5, 0.5, 0.6, 1.0, 0.4, 0.4, 0.7, 0.8, 0.7)) |>
  fontsize(size = 9, part = "all") |> # Fonte menor para tabelas densas
  align(align = "center", part = "header") |>
  align(align = "left", part = "body") |>
  bold(part = "header") |>
  valign(valign = "top", part = "body") |>
  set_caption(caption = vcap)

# Visualizar
ft_desc
```

## Forest Cover Map from MapBiomas6

The processed forest cover map from MapBiomas collection 6 used in the simulations were displayed in figure SI3.5 e SI3.6.

# Geoprocessing

## Example of plot relocation to the immediate neighborhood   {.unnumbered}

```{r deslocamento parcela, eval=FALSE, include=FALSE}
library(purrr)
library(magick)
library(cowplot)
library(ggplot2)
img_to_gg <- function(img) {
  ggdraw() + 
    draw_image(img) +  # Scale de 0.9 para margens
    theme_void()
}
img <- image_read("0_apendices/mapas_p_MNEE/figuras/exemplo_redesenho_parcela.png") %>% 
  image_trim()
img_crop <- image_crop(img, "558x475+0+16")
plot <- img_to_gg(img_crop)
saveRDS(plot,"1_to_compile_dissertacao_EM_USO/09_SI/RDS/deslocamento_parcela.rds")
```
```{r caption deslocamento das parcelas}
vcap <- "Example of plot displacement to the immediate neighborhood. Due to the spatial configuration of the forest cover, some plots could not be centered within the landscape. In these cases, the plot was relocated to the nearest available forest area within the same forest patch. In the figure, dark red represents the adjusted sampled plot area, yellow represents the remaining forest cover, and white represents the non-forest matrix."
```
```{r desloc-parcela, fig.height=2,fig.width=4.5,fig.cap="Exemplo de deslocamenteo da parcela quadrada do centro da parcela para a vizinhança próxima."}
plot <- readRDS("1_to_compile_dissertacao_EM_USO/09_SI/RDS/deslocamento_parcela.rds")
plot
```

## Forest Fragments with Insufficient Area for Square Plots

```{r criacao sitios remov, eval=FALSE, include=FALSE}
limg <- c("0_apendices/mapas_p_MNEE/figuras/sitio_removido.png",
          "0_apendices/mapas_p_MNEE/figuras/sitio_removido_BAURUC.png")
limg <- lapply(limg,img_to_gg)
saveRDS(limg,file="1_to_compile_dissertacao_EM_USO/09_SI/RDS/mapas_semquadrado.rds")
```
```{r sitios-removidos, fig.cap="Detail of two forest fragments where it was not possible to allocate a square plot due to limited continuous habitat. The cutouts show that although there is forest in the surroundings, there is insufficient area to shift the plot to another position within the same fragment without exceeding the boundaries of the target habitat."}
library(gridExtra)
p <- readRDS(file="1_to_compile_dissertacao_EM_USO/09_SI/RDS/mapas_semquadrado.rds")
grid.arrange(grobs=p,nrow=1)
```

## Exemple of the experimental landscape

```{r caption fig exemplo paisgens experimentais}
vcap <- "Comparison of landscape spatial configurations for the neutral model simulations. For each forest plot, three landscape scenarios were generated based on the observed total forest cover: (left) Fragmented, representing the real-world spatial configuration derived from MapBiomas; (center) Clumped, where the total forest cover is preserved but aggregated into a single, continuous, and centered patch; and (right) Pristine, a baseline scenario where the entire landscape is assumed to be fully forested (100% forest cover). Colors follow a terrain-based scale where dark green represents the sampled plot area, yellow represents the remaining forest cover, and white represents the non-forest matrix. These scenarios allow for the decomposition of landscape effects into total forest cover loss (Pristine vs. Clumped) and fragmentation per se (Clumped vs. Fragmented)."
```
```{r fig-exemplo-paisagens-experimentais, fig.width=9, fig.height=3.5, fig.cap=vcap}
source("source/mapa_p_MNEE.R")
f_image <- \(mat,title){
  image(mat,
        main=title,
        zlim=c(0,2),
        axes=FALSE,
        col=terrain.colors(12,rev = TRUE))
}
#
par(mfrow = c(1, 3), mar = c(1, 1, 3, 1)) # Diminuir margens externas ajuda muito
#
path_name <- "dados/simulacao/MGipia1.txt"
m <- read_table(file = path_name) |> as.matrix()
f_image(mat=m,title="Fragmented")
# sem frag
m_sem_frag <- f_nonFragLand(m)
f_image(mat=m_sem_frag,title="Clumped")
# idealizado
m_ideal <- m
m_ideal[m_ideal==0] <- 1
f_image(mat=m_ideal,title="Pristine")
```


# Dispersal kernel

<p>
Dispersal distances were modeled using a Laplace kernel, which captures the canonical pattern expected for dispersal processes—high propagule density near the source and decreasing probability with distance. 
Although no single statistical distribution universally describes real dispersal kernels, exponential-tailed functions provide a tractable and commonly used approximation of dispersal (REF 2017). In this study, the Laplace kernel was adopted as a parsimonious representation of neighbourhood influence, suitable for a first approximation of connectivity in the spatially explicit neutral simulations. 
This formulation allowed us to explore uncertainty in dispersal limitation while maintaining a mechanistically interpretable representation of neighbourhood recruitment.
</p>

<p>
Dispersal distances were sampled from a Laplace distribution. The probability density function is given by
</p>


$$ f(x) = \frac{1}{2b} \exp\left(-\frac{|x|}{b}\right) $$

The dispersal kernel was parameterized using the scale parameter b. In our implementation, the distribution was defined in terms of its standard deviation ($\sigma$), such that:

$$ \sigma = \sqrt{2} \cdot b $$
and therefore

$$ b = \frac{\sigma}{\sqrt{2}} $$
Random dispersal distances were generated using the inverse cumulative distribution function (inverse-CDF) method. All simulations were implemented in C++ following the implementation described by James Rosindell et al. (2008; Appendix X).

## Empirical Calibration of the dispersal kernel

For each plot, the parameter b was calibrated according to the local tree density and the predefined dispersal scenario (k). The parameter k represents the fraction of propagules that disperse locally (i.e., within the area of the parent and its immediate neighbors). The explored scenarios range from k = 0.99, representing severe dispersal limitation, to k = 0.05, representing high dispersal capability. As expected, calibrated b values increase as k and density decreases (Fig. \@ref(fig:fig-kcalib)), reflecting the larger mean dispersal distances required to achieve lower local retention.

```{r caption k calibration,eval=TRUE,include=TRUE,echo=FALSE}
vcap <- "Empirical calibration of the dispersal scale parameter (b) across plots and dispersal scenarios (k). Each boxplot shows the distribution of calibrated b values for all forest plots under different dispersal scenarios (k). The parameter k represents the fraction of propagules dispersing locally, ranging from 0.99 (severe dispersal limitation) to 0.05 (high dispersal capability). For each scenario, b was calibrated based on the plot-specific tree density (individuals /plot area), as shown by the color gradient. The observed increase in b at lower k values reflects the larger standard deviation required to achieve greater dispersal distances. Each point represents an individual forest plot (n = 105)."
```
```{r fig-kcalib,fig.cap=vcap,fig.width=8,include=TRUE,eval=TRUE,echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
# df dados da proporção de cobertura vegetal
df_p <- read_csv("dados/df_p.csv")
# df dados para simulação
df_sim <- read_csv("dados/df_simulacao.csv") |> 
  inner_join(x=df_p,by="SiteCode")
df_sim |> 
  mutate(DA = Ntotal / effort_ha) %>% 
  ggplot(aes(x=k,y=d,group=k,color=DA)) +
  geom_boxplot() +
  geom_jitter(alpha=0.8) +
  scale_x_continuous(breaks = unique(df_sim$k)) +
  scale_color_viridis_c(name = expression(Density~(ind.ha^-1))) +
  labs(x = "Dispersal scenario (k)",
       y = "Calibrated parameter (b)") +
  theme_bw()
```

